---
title: "CP34H (*Colwellia psychrerythraea* 34H) — Supplemental BC vs. E Analyses"
subtitle: "Control vs. DNA Shield 1:1.5 | NSAF density scatter · NSAF proportions test · PSM proportions test"
author: "Evelyn Bartlett (Brook Lab)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7.6,
  fig.height = 7,
  dpi = 300
)
```

# Overview

Three supplemental analyses comparing the **bacterial control (BC, n = 3)** against
the **DNA Shield 1:1.5 treatment (E, n = 3)** for *Colwellia psychrerythraea* 34H:

1. **Section 1 — NSAF Density Scatter:** Per-protein log₁₀ NSAF scatter with 2D
   density colouring and ± SD error bars.
2. **Section 2 — NSAF Proportions Test:** Per-protein two-sample test of proportions
   on NSAF values. p-values are Benjamini–Hochberg adjusted.
3. **Section 3 — PSM Proportions Test:** Per-protein two-sample test of proportions
   on raw PSM counts. p-values are Benjamini–Hochberg adjusted.

Sections 2 and 3 each analyse two input files separately:

- `Bactera_Comet_all_samples_proteins.txt` — full-elution table
- `Bacteria_Comet_elution_3_to_24min_prot.txt` — restricted to the 3–24 min window

Upstream pipeline: Comet 2023.02 via Limelight, PSM- and peptide-level q ≤ 0.01,
no protein-level FDR applied. **HR-E1 is a known PCA outlier** — included for
completeness; consider re-running with HR-E1 dropped if it drives false positives.

---

```{r packages}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(viridis)  # density scatter colour scale
library(MASS)     # kde2d for 2D density
```

---

# Section 1 — NSAF Density Scatter (BC vs. E)

## Overview

Per-protein scatter of mean NSAF (control BC vs. Shield 1:1.5 E). Both axes
are log₁₀. Point colour = local 2D density (viridis: purple = sparse, yellow =
dense). Whiskers = ± SD across biological triplicates, clipped to the axis floor.
Dashed grey line = 1:1 (no change).

## Load data

```{r ds_load}
data_file <- "Bacteria/Bactera_Comet_all_samples_proteins.txt"

raw_ds <- read_tsv(
  data_file,
  show_col_types = FALSE,
  na = c("", "NA")
)

# Column names mix '_' and '-' (e.g. "NSAF (HR-E1)" vs "NSAF (HR_E2)")
nsaf_bc <- c("NSAF (HR_BC1)", "NSAF (HR_BC2)", "NSAF (HR_BC3)")
nsaf_e  <- c("NSAF (HR-E1)",  "NSAF (HR_E2)",  "NSAF (HR_E3)")

stopifnot(all(c(nsaf_bc, nsaf_e) %in% colnames(raw_ds)))
```

## Compute per-protein mean and SD

```{r ds_summarise}
prot <- raw_ds %>%
  mutate(across(all_of(c(nsaf_bc, nsaf_e)), as.numeric)) %>%
  rowwise() %>%
  mutate(
    BC_mean = mean(c_across(all_of(nsaf_bc)), na.rm = TRUE),
    BC_sd   = sd(  c_across(all_of(nsaf_bc)), na.rm = TRUE),
    E_mean  = mean(c_across(all_of(nsaf_e)),  na.rm = TRUE),
    E_sd    = sd(  c_across(all_of(nsaf_e)),  na.rm = TRUE)
  ) %>%
  ungroup() %>%
  dplyr::select(`Protein(s)`, `Protein Description(s)`,
                BC_mean, BC_sd, E_mean, E_sd) %>%
  mutate(BC_sd = tidyr::replace_na(BC_sd, 0),
         E_sd  = tidyr::replace_na(E_sd,  0))

# Keep proteins detected in at least one condition
plot_df <- prot %>% filter(BC_mean > 0 | E_mean > 0)
cat("Proteins plotted:", nrow(plot_df), "\n")
```

## Set up log axes and clip error bars

```{r ds_limits}
pos_vals <- c(plot_df$BC_mean, plot_df$E_mean)
pos_vals <- pos_vals[pos_vals > 0]
ax_lo <- max(min(pos_vals) * 0.5, 1e-7)
ax_hi <- max(pos_vals) * 2

plot_df <- plot_df %>%
  mutate(
    BC_plot = pmax(BC_mean, ax_lo),
    E_plot  = pmax(E_mean,  ax_lo),
    BC_lo   = pmax(BC_mean - BC_sd, ax_lo),
    BC_hi   = pmax(BC_mean + BC_sd, BC_plot),
    E_lo    = pmax(E_mean  - E_sd,  ax_lo),
    E_hi    = pmax(E_mean  + E_sd,  E_plot)
  )
```

## Compute 2D density in log space

```{r ds_density}
get_log_point_density <- function(x, y, n = 120) {
  lx <- log10(x); ly <- log10(y)
  dens <- MASS::kde2d(lx, ly, n = n)
  ix <- findInterval(lx, dens$x); iy <- findInterval(ly, dens$y)
  ix <- pmin(pmax(ix, 1), n); iy <- pmin(pmax(iy, 1), n)
  dens$z[cbind(ix, iy)]
}

plot_df <- plot_df %>%
  mutate(
    dens     = get_log_point_density(BC_plot, E_plot, n = 120),
    dens_log = log10(dens + 1)
  ) %>%
  arrange(dens_log)
```

## Density scatter plot

```{r ds_plot, fig.width=7.6, fig.height=7}
p_density <- ggplot(plot_df, aes(x = BC_plot, y = E_plot)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey40") +
  geom_errorbarh(aes(xmin = BC_lo, xmax = BC_hi),
                 height = 0, colour = "grey55",
                 linewidth = 0.25, alpha = 0.30) +
  geom_errorbar(aes(ymin = E_lo, ymax = E_hi),
                width = 0, colour = "grey55",
                linewidth = 0.25, alpha = 0.30) +
  geom_point(aes(colour = dens_log),
             size = 1.2, alpha = 0.9) +
  scale_colour_viridis_c(option = "viridis",
                         name = expression(log[10](density + 1))) +
  scale_x_log10(limits = c(ax_lo, ax_hi),
                labels = trans_format("log10", math_format(10^.x))) +
  scale_y_log10(limits = c(ax_lo, ax_hi),
                labels = trans_format("log10", math_format(10^.x))) +
  coord_equal() +
  labs(
    x = "Mean NSAF — BC (control)",
    y = "Mean NSAF — E (DNA Shield 1:1.5)",
    title = "Protein abundance: Control (BC) vs. DNA Shield 1:1.5 (E)",
    subtitle = sprintf(
      "CP34H (Colwellia psychrerythraea 34H) — n = %d proteins (log–log, purple = sparse, yellow = dense)",
      nrow(plot_df)
    )
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey30"),
    legend.position = "right"
  )

p_density
```

## Save — Section 1

```{r ds_save}
out_ds <- "figures/NSAF_BC_vs_E_density_scatter"
dir.create(out_ds, recursive = TRUE, showWarnings = FALSE)

ggsave(file.path(out_ds, "bacteria_BC_vs_E_NSAF_density_log.png"),
       p_density, width = 7.6, height = 7, dpi = 300)
ggsave(file.path(out_ds, "bacteria_BC_vs_E_NSAF_density_log.pdf"),
       p_density, width = 7.6, height = 7)

readr::write_tsv(
  plot_df %>% dplyr::select(`Protein(s)`, `Protein Description(s)`,
                            BC_mean, BC_sd, E_mean, E_sd),
  file.path(out_ds, "bacteria_BC_vs_E_NSAF_summary.tsv")
)

message("Section 1 outputs written to: ", out_ds)
```

---

# Section 2 — NSAF Proportions Test (BC vs. E)

## Overview

Per-protein two-sample test of proportions on NSAF values. NSAF is already a
within-sample proportion, so replicate NSAFs are summed per group and scaled by
`SCALE = 1e6` to produce count-like integers for `prop.test()`. Fisher's exact
test is used as a fallback when any expected cell < 5. p-values are
Benjamini–Hochberg adjusted. Two files are analysed separately.

## Helper function

```{r nsaf_helper}
nsaf_bc_cols <- c("NSAF (HR_BC1)", "NSAF (HR_BC2)", "NSAF (HR_BC3)")
nsaf_e_cols  <- c("NSAF (HR-E1)",  "NSAF (HR_E2)",  "NSAF (HR_E3)")

parse_nsaf <- function(x) {
  as.numeric(gsub(",", "", as.character(x), fixed = TRUE))
}

SCALE <- 1e6

run_nsaf_prop_test <- function(file_path, label) {
  raw <- read_tsv(file_path, show_col_types = FALSE, na = c("", "NA"))
  stopifnot(all(c(nsaf_bc_cols, nsaf_e_cols) %in% colnames(raw)))

  dat <- raw %>%
    mutate(across(all_of(c(nsaf_bc_cols, nsaf_e_cols)), parse_nsaf)) %>%
    rowwise() %>%
    mutate(
      sum_BC  = sum(c_across(all_of(nsaf_bc_cols)), na.rm = TRUE),
      sum_E   = sum(c_across(all_of(nsaf_e_cols)),  na.rm = TRUE),
      mean_BC = mean(c_across(all_of(nsaf_bc_cols)), na.rm = TRUE),
      mean_E  = mean(c_across(all_of(nsaf_e_cols)),  na.rm = TRUE)
    ) %>%
    ungroup() %>%
    dplyr::select(`Protein(s)`, `Protein Description(s)`,
                  sum_BC, sum_E, mean_BC, mean_E) %>%
    mutate(
      k_BC = as.integer(round(sum_BC * SCALE)),
      k_E  = as.integer(round(sum_E  * SCALE))
    )

  N_BC <- sum(dat$k_BC, na.rm = TRUE)
  N_E  <- sum(dat$k_E,  na.rm = TRUE)
  cat(sprintf("[%s] proteins: %d   N_BC=%d   N_E=%d   (SCALE=%.0e)\n",
              label, nrow(dat), N_BC, N_E, SCALE))

  dat <- dat %>% filter(!(k_BC == 0 & k_E == 0))

  one_test <- function(k_BC, k_E, N_BC, N_E) {
    tab <- matrix(c(k_BC, N_BC - k_BC, k_E, N_E - k_E), nrow = 2, byrow = TRUE)
    expected <- outer(rowSums(tab), colSums(tab)) / sum(tab)
    if (any(expected < 5)) {
      ft <- suppressWarnings(fisher.test(tab))
      list(p = ft$p.value, method = "Fisher")
    } else {
      pt <- suppressWarnings(prop.test(x = c(k_BC, k_E), n = c(N_BC, N_E), correct = TRUE))
      list(p = pt$p.value, method = "prop.test")
    }
  }

  eps <- 0.5 / SCALE
  results <- dat %>%
    rowwise() %>%
    mutate(
      .res    = list(one_test(k_BC, k_E, N_BC, N_E)),
      p_value = .res$p,
      method  = .res$method
    ) %>%
    ungroup() %>%
    dplyr::select(-.res) %>%
    mutate(
      prop_BC   = sum_BC / (N_BC / SCALE),
      prop_E    = sum_E  / (N_E  / SCALE),
      log2FC    = log2(mean_E + eps) - log2(mean_BC + eps),
      q_value   = p.adjust(p_value, method = "BH"),
      direction = case_when(
        q_value >= 0.05 ~ "ns",
        log2FC  >  0    ~ "up in E (1:1.5)",
        log2FC  <  0    ~ "down in E (1:1.5)",
        TRUE            ~ "ns"
      ),
      source_file = label
    ) %>%
    arrange(q_value, p_value)

  results
}
```

## Run — all-samples file

```{r nsaf_file1}
nsaf_res_all <- run_nsaf_prop_test(
  "Bacteria/Bactera_Comet_all_samples_proteins.txt",
  label = "all_samples"
)

nsaf_res_all %>%
  summarise(n_proteins = n(), n_sig_q05 = sum(q_value < 0.05),
            n_up_q05 = sum(q_value < 0.05 & log2FC > 0),
            n_down_q05 = sum(q_value < 0.05 & log2FC < 0))
```

### Top 20 — all-samples

```{r nsaf_file1_top}
nsaf_res_all %>%
  dplyr::select(`Protein(s)`, `Protein Description(s)`,
                mean_BC, mean_E, log2FC, method, p_value, q_value, direction) %>%
  head(20) %>%
  knitr::kable(digits = c(NA, NA, 5, 5, 3, NA, 4, 4, NA))
```

## Run — elution 3–24 min file

```{r nsaf_file2}
nsaf_res_elution <- run_nsaf_prop_test(
  "Bacteria/Bacteria_Comet_elution_3_to_24min_prot.txt",
  label = "elution_3_to_24min"
)

nsaf_res_elution %>%
  summarise(n_proteins = n(), n_sig_q05 = sum(q_value < 0.05),
            n_up_q05 = sum(q_value < 0.05 & log2FC > 0),
            n_down_q05 = sum(q_value < 0.05 & log2FC < 0))
```

### Top 20 — elution 3–24 min

```{r nsaf_file2_top}
nsaf_res_elution %>%
  dplyr::select(`Protein(s)`, `Protein Description(s)`,
                mean_BC, mean_E, log2FC, method, p_value, q_value, direction) %>%
  head(20) %>%
  knitr::kable(digits = c(NA, NA, 5, 5, 3, NA, 4, 4, NA))
```

## Volcano plots — NSAF proportions

```{r nsaf_volcano}
make_volcano_nsaf <- function(df, title_tag) {
  ggplot(df, aes(x = log2FC, y = -log10(p_value), colour = direction)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
    geom_hline(yintercept = -log10(0.05), linetype = "dotted", colour = "grey50", linewidth = 0.4) +
    geom_point(alpha = 0.7, size = 1.4) +
    scale_colour_manual(
      values = c("ns" = "grey70", "up in E (1:1.5)" = "#1f78b4", "down in E (1:1.5)" = "#08306b"),
      name = "BH q < 0.05"
    ) +
    labs(
      x = expression(log[2]~"(mean NSAF, E / BC)"),
      y = expression(-log[10]~"(p)"),
      title    = paste0("NSAF proportions test — ", title_tag),
      subtitle = sprintf("CP34H · n = %d proteins · %d significant at q<0.05",
                         nrow(df), sum(df$q_value < 0.05))
    ) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(colour = "grey30", size = 9))
}

nsaf_v_all     <- make_volcano_nsaf(nsaf_res_all,     "all-samples")
nsaf_v_elution <- make_volcano_nsaf(nsaf_res_elution, "elution 3–24 min")

nsaf_v_all
nsaf_v_elution
```

## Save — Section 2

```{r nsaf_save}
out_nsaf <- "figures/NSAF_proportions_test_BC_vs_E"
dir.create(out_nsaf, recursive = TRUE, showWarnings = FALSE)

readr::write_tsv(nsaf_res_all,
  file.path(out_nsaf, "bacteria_BC_vs_E_NSAF_proptest_all_samples.tsv"))
readr::write_tsv(nsaf_res_elution,
  file.path(out_nsaf, "bacteria_BC_vs_E_NSAF_proptest_elution_3_to_24min.tsv"))

ggsave(file.path(out_nsaf, "volcano_NSAF_BC_vs_E_all_samples.png"),
       nsaf_v_all, width = 7.5, height = 6, dpi = 600, bg = "white")
ggsave(file.path(out_nsaf, "volcano_NSAF_BC_vs_E_all_samples.pdf"),
       nsaf_v_all, width = 7.5, height = 6)
ggsave(file.path(out_nsaf, "volcano_NSAF_BC_vs_E_elution_3_to_24min.png"),
       nsaf_v_elution, width = 7.5, height = 6, dpi = 600, bg = "white")
ggsave(file.path(out_nsaf, "volcano_NSAF_BC_vs_E_elution_3_to_24min.pdf"),
       nsaf_v_elution, width = 7.5, height = 6)

message("Section 2 outputs written to: ", out_nsaf)
```

---

# Section 3 — PSM Proportions Test (BC vs. E)

## Overview

Per-protein two-sample test of proportions on raw PSM counts. For each protein,
replicate PSMs are summed within group to form a 2×2 contingency table (protein
PSMs vs. all other PSMs). Chi-squared via `prop.test()` when expected counts ≥ 5,
Fisher's exact test otherwise. p-values are Benjamini–Hochberg adjusted.

## Helper function

```{r psm_helper}
psm_bc_cols <- c("PSMs (HR_BC1)", "PSMs (HR_BC2)", "PSMs (HR_BC3)")
psm_e_cols  <- c("PSMs (HR-E1)",  "PSMs (HR_E2)",  "PSMs (HR_E3)")

parse_psm <- function(x) {
  as.integer(gsub(",", "", as.character(x), fixed = TRUE))
}

run_psm_prop_test <- function(file_path, label) {
  raw <- read_tsv(file_path, show_col_types = FALSE, na = c("", "NA"))
  stopifnot(all(c(psm_bc_cols, psm_e_cols) %in% colnames(raw)))

  dat <- raw %>%
    mutate(across(all_of(c(psm_bc_cols, psm_e_cols)), parse_psm)) %>%
    rowwise() %>%
    mutate(
      k_BC = sum(c_across(all_of(psm_bc_cols)), na.rm = TRUE),
      k_E  = sum(c_across(all_of(psm_e_cols)),  na.rm = TRUE)
    ) %>%
    ungroup() %>%
    dplyr::select(`Protein(s)`, `Protein Description(s)`, k_BC, k_E)

  N_BC <- sum(dat$k_BC, na.rm = TRUE)
  N_E  <- sum(dat$k_E,  na.rm = TRUE)
  cat(sprintf("[%s] proteins: %d   N_BC=%d   N_E=%d\n", label, nrow(dat), N_BC, N_E))

  dat <- dat %>% filter(!(k_BC == 0 & k_E == 0))

  one_test <- function(k_BC, k_E, N_BC, N_E) {
    tab <- matrix(c(k_BC, N_BC - k_BC, k_E, N_E - k_E), nrow = 2, byrow = TRUE)
    expected <- outer(rowSums(tab), colSums(tab)) / sum(tab)
    if (any(expected < 5)) {
      ft <- suppressWarnings(fisher.test(tab))
      list(p = ft$p.value, method = "Fisher")
    } else {
      pt <- suppressWarnings(prop.test(x = c(k_BC, k_E), n = c(N_BC, N_E), correct = TRUE))
      list(p = pt$p.value, method = "prop.test")
    }
  }

  results <- dat %>%
    rowwise() %>%
    mutate(
      .res    = list(one_test(k_BC, k_E, N_BC, N_E)),
      p_value = .res$p,
      method  = .res$method
    ) %>%
    ungroup() %>%
    dplyr::select(-.res) %>%
    mutate(
      prop_BC   = k_BC / N_BC,
      prop_E    = k_E  / N_E,
      log2FC    = log2((k_E + 1) / (N_E + 1)) - log2((k_BC + 1) / (N_BC + 1)),
      q_value   = p.adjust(p_value, method = "BH"),
      direction = case_when(
        q_value >= 0.05 ~ "ns",
        log2FC  >  0    ~ "up in E (1:1.5)",
        log2FC  <  0    ~ "down in E (1:1.5)",
        TRUE            ~ "ns"
      ),
      source_file = label
    ) %>%
    arrange(q_value, p_value)

  results
}
```

## Run — all-samples file

```{r psm_file1}
psm_res_all <- run_psm_prop_test(
  "Bacteria/Bactera_Comet_all_samples_proteins.txt",
  label = "all_samples"
)

psm_res_all %>%
  summarise(n_proteins = n(), n_sig_q05 = sum(q_value < 0.05),
            n_up_q05 = sum(q_value < 0.05 & log2FC > 0),
            n_down_q05 = sum(q_value < 0.05 & log2FC < 0))
```

### Top 20 — all-samples

```{r psm_file1_top}
psm_res_all %>%
  dplyr::select(`Protein(s)`, `Protein Description(s)`,
                k_BC, k_E, prop_BC, prop_E, log2FC, method, p_value, q_value, direction) %>%
  head(20) %>%
  knitr::kable(digits = c(NA, NA, 0, 0, 5, 5, 3, NA, 4, 4, NA))
```

## Run — elution 3–24 min file

```{r psm_file2}
psm_res_elution <- run_psm_prop_test(
  "Bacteria/Bacteria_Comet_elution_3_to_24min_prot.txt",
  label = "elution_3_to_24min"
)

psm_res_elution %>%
  summarise(n_proteins = n(), n_sig_q05 = sum(q_value < 0.05),
            n_up_q05 = sum(q_value < 0.05 & log2FC > 0),
            n_down_q05 = sum(q_value < 0.05 & log2FC < 0))
```

### Top 20 — elution 3–24 min

```{r psm_file2_top}
psm_res_elution %>%
  dplyr::select(`Protein(s)`, `Protein Description(s)`,
                k_BC, k_E, prop_BC, prop_E, log2FC, method, p_value, q_value, direction) %>%
  head(20) %>%
  knitr::kable(digits = c(NA, NA, 0, 0, 5, 5, 3, NA, 4, 4, NA))
```

## Volcano plots — PSM proportions

```{r psm_volcano}
make_volcano_psm <- function(df, title_tag) {
  ggplot(df, aes(x = log2FC, y = -log10(p_value), colour = direction)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
    geom_hline(yintercept = -log10(0.05), linetype = "dotted", colour = "grey50", linewidth = 0.4) +
    geom_point(alpha = 0.7, size = 1.4) +
    scale_colour_manual(
      values = c("ns" = "grey70", "up in E (1:1.5)" = "#1f78b4", "down in E (1:1.5)" = "#08306b"),
      name = "BH q < 0.05"
    ) +
    labs(
      x = expression(log[2]~"(E / BC)"),
      y = expression(-log[10]~"(p)"),
      title    = paste0("PSM proportions test — ", title_tag),
      subtitle = sprintf("CP34H · n = %d proteins · %d significant at q<0.05",
                         nrow(df), sum(df$q_value < 0.05))
    ) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(colour = "grey30", size = 9))
}

psm_v_all     <- make_volcano_psm(psm_res_all,     "all-samples")
psm_v_elution <- make_volcano_psm(psm_res_elution, "elution 3–24 min")

psm_v_all
psm_v_elution
```

## Save — Section 3

```{r psm_save}
out_psm <- "figures/PSM_proportions_test_BC_vs_E"
dir.create(out_psm, recursive = TRUE, showWarnings = FALSE)

readr::write_tsv(psm_res_all,
  file.path(out_psm, "bacteria_BC_vs_E_PSM_proptest_all_samples.tsv"))
readr::write_tsv(psm_res_elution,
  file.path(out_psm, "bacteria_BC_vs_E_PSM_proptest_elution_3_to_24min.tsv"))

ggsave(file.path(out_psm, "volcano_PSM_BC_vs_E_all_samples.png"),
       psm_v_all, width = 7.5, height = 6, dpi = 600, bg = "white")
ggsave(file.path(out_psm, "volcano_PSM_BC_vs_E_all_samples.pdf"),
       psm_v_all, width = 7.5, height = 6)
ggsave(file.path(out_psm, "volcano_PSM_BC_vs_E_elution_3_to_24min.png"),
       psm_v_elution, width = 7.5, height = 6, dpi = 600, bg = "white")
ggsave(file.path(out_psm, "volcano_PSM_BC_vs_E_elution_3_to_24min.pdf"),
       psm_v_elution, width = 7.5, height = 6)

message("Section 3 outputs written to: ", out_psm)
```

---

# Session Info

```{r session}
sessionInfo()
```
