---
title: "DNA Shield — CP34H (*Colwellia psychrerythraea* 34H) Differential Protein Abundance"
subtitle: "DDA Proteomics | log2(NSAF) + Median Centering + limma eBayes"
author: "Evelyn Bartlett — Brook Lab"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 6,
  dpi     = 300
)
```

# Overview

This document performs differential protein abundance analysis for the DNA Shield
project bacterial arm (CP34H (*Colwellia psychrerythraea* 34H)). Samples are divided into:

- **Control:** BC1, BC2, BC3
- **Experimental:** E1, E2, E3

**Pipeline:** filter → log2(NSAF) → median centering → limma moderated t-test
(Smyth 2004) → Benjamini-Hochberg FDR correction.

---

# 1 · Load packages

```{r packages}
# Install any missing packages before running:
# install.packages(c("tidyverse", "limma", "pheatmap", "ggrepel",
#                    "RColorBrewer", "viridis", "scales", "openxlsx"))
# BiocManager::install("limma")

library(tidyverse)
library(limma)
library(pheatmap)
library(ggrepel)
library(RColorBrewer)
library(viridis)
library(scales)
library(openxlsx)
```

---

# 2 · Load data

```{r load-data}
# ── Update this path to match your local file location ──────────────────────
DATA_PATH <- "Bactera_Comet_all_samples_proteins.txt"

raw <- read.delim(DATA_PATH, stringsAsFactors = FALSE, check.names = FALSE)
cat("Raw protein rows:", nrow(raw), "\n")
```

---

# 3 · Define sample groups

```{r sample-cols}
# PSM (spectral count) columns
psm_ctrl <- c("PSMs (HR_BC1)", "PSMs (HR_BC2)", "PSMs (HR_BC3)")
psm_exp  <- c("PSMs (HR-E1)",  "PSMs (HR_E2)",  "PSMs (HR_E3)")

# NSAF columns
nsaf_ctrl <- c("NSAF (HR_BC1)", "NSAF (HR_BC2)", "NSAF (HR_BC3)")
nsaf_exp  <- c("NSAF (HR-E1)",  "NSAF (HR_E2)",  "NSAF (HR_E3)")

all_psm  <- c(psm_ctrl,  psm_exp)
all_nsaf <- c(nsaf_ctrl, nsaf_exp)

# Clean commas from numeric fields (Limelight/Comet export format)
raw[all_psm]  <- lapply(raw[all_psm],  function(x) as.numeric(gsub(",", "", x)))
raw[all_nsaf] <- lapply(raw[all_nsaf], function(x) as.numeric(gsub(",", "", x)))
```

---

# 4 · Filter proteins

## 4.1 Remove contaminants

```{r filter-contaminants}
contam_mask <- grepl("\\(CONTAMINANT\\)", raw[["Protein Description(s)"]], ignore.case = TRUE)
cat("Contaminant rows removed:", sum(contam_mask), "\n")
dat <- raw[!contam_mask, ]
```

## 4.2 Remove *Thalassiosira pseudonana* proteins

```{r filter-thalassiosira}
thala_mask <- grepl("Thalassiosira", dat[["Protein Description(s)"]], ignore.case = TRUE) |
              grepl("Thalassiosira", dat[["Protein(s)"]],             ignore.case = TRUE)
cat("Thalassiosira pseudonana rows removed:", sum(thala_mask), "\n")
dat <- dat[!thala_mask, ]
cat("Rows after organism filters:", nrow(dat), "\n")
```

## 4.3 Spectral count quality filter

Retain proteins with **≥ 2 spectral counts in ≥ 2 of 3 replicates** in at
least one condition.

```{r filter-spc}
passes_spc <- function(df, cols) {
  rowSums(df[cols] >= 2) >= 2
}

ctrl_pass <- passes_spc(dat, psm_ctrl)
exp_pass  <- passes_spc(dat, psm_exp)
dat <- dat[ctrl_pass | exp_pass, ]
dat <- dat[!duplicated(dat[["Protein Group Number"]]), ]  # one row per group
dat <- dat %>% mutate(protein_id = paste0("pg", `Protein Group Number`))
rownames(dat) <- dat$protein_id

cat("Proteins after SpC filter:", nrow(dat), "\n")
```

---

# 5 · Normalisation

## 5.1 log2(NSAF + pseudocount)

```{r log2-transform}
PSEUDO <- 1e-8
nsaf_mat <- as.matrix(dat[, all_nsaf])
log2_mat <- log2(nsaf_mat + PSEUDO)
colnames(log2_mat) <- c("BC1", "BC2", "BC3", "E1", "E2", "E3")
rownames(log2_mat) <- dat$protein_id
```

## 5.2 Median centering

```{r median-center}
col_medians <- apply(log2_mat, 2, median)
log2_norm   <- sweep(log2_mat, 2, col_medians, "-")

cat("Sample medians subtracted:\n")
print(round(col_medians, 4))
```

## 5.3 QC: NSAF distribution before and after normalisation

```{r qc-boxplot-raw, fig.cap="Fig 1a. log2(NSAF) distribution before median centering."}
# Project colour palette — bacteria arm uses blue family
sample_cols <- c(BC1 = "#2166ac", BC2 = "#4393c3", BC3 = "#92c5de",
                 E1  = "#d6604d", E2  = "#f4a582", E3  = "#fddbc7")

df_raw <- as.data.frame(log2_mat) %>%
  rownames_to_column("protein") %>%
  pivot_longer(-protein, names_to = "sample", values_to = "log2NSAF")

ggplot(df_raw, aes(x = sample, y = log2NSAF, fill = sample)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.85) +
  scale_fill_manual(values = sample_cols) +
  labs(title = "log2(NSAF) — before median centering",
       x = NULL, y = "log2(NSAF)") +
  theme_bw(base_size = 13) +
  theme(legend.position = "none")
```

```{r qc-boxplot-norm, fig.cap="Fig 1b. log2(NSAF) distribution after median centering."}
df_norm <- as.data.frame(log2_norm) %>%
  rownames_to_column("protein") %>%
  pivot_longer(-protein, names_to = "sample", values_to = "log2NSAF")

ggplot(df_norm, aes(x = sample, y = log2NSAF, fill = sample)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.85) +
  scale_fill_manual(values = sample_cols) +
  labs(title = "log2(NSAF) — after median centering",
       x = NULL, y = "log2(NSAF)") +
  theme_bw(base_size = 13) +
  theme(legend.position = "none")
```

---

# 6 · limma moderated t-test

```{r limma}
# Design matrix: intercept = Control, coef 2 = Experimental effect
group       <- factor(c("Ctrl","Ctrl","Ctrl","Exp","Exp","Exp"),
                      levels = c("Ctrl","Exp"))
design      <- model.matrix(~group)
colnames(design) <- c("Intercept", "Exp_vs_Ctrl")

fit         <- lmFit(log2_norm, design)
fit_eb      <- eBayes(fit)

results_tbl <- topTable(fit_eb,
                         coef     = "Exp_vs_Ctrl",
                         number   = Inf,
                         sort.by  = "p",
                         adjust.method = "BH") %>%
  as_tibble(rownames = "protein_id") %>%
  left_join(dat %>% select(protein_id,
                            `Protein(s)`,
                            `Protein Description(s)`),
            by = "protein_id") %>%
  rename(logFC        = logFC,
         mean_expr    = AveExpr,
         mod_t        = t,
         p_value      = P.Value,
         adj_p_BH     = adj.P.Val,
         B_statistic  = B)

cat(sprintf("Prior df (d0):        %.2f\n", fit_eb$df.prior))
cat(sprintf("Prior variance (s0²): %.6f\n", fit_eb$s2.prior))
cat(sprintf("Posterior df:         %.2f\n", fit_eb$df.prior + fit_eb$df.residual[1]))
cat(sprintf("\nProteins tested:         %d\n", nrow(results_tbl)))
cat(sprintf("Significant FDR < 0.05:  %d (up: %d, down: %d)\n",
    sum(results_tbl$adj_p_BH < 0.05),
    sum(results_tbl$adj_p_BH < 0.05 & results_tbl$logFC > 0),
    sum(results_tbl$adj_p_BH < 0.05 & results_tbl$logFC < 0)))
cat(sprintf("Significant FDR < 0.01:  %d\n",
    sum(results_tbl$adj_p_BH < 0.01)))
```

---

# 7 · Figures

## 7.1 PCA — sample clustering

```{r pca, fig.cap="Fig 2. PCA of all 6 samples on median-centred log2(NSAF)."}
pca_res  <- prcomp(t(log2_norm), scale. = FALSE)
pca_df   <- as.data.frame(pca_res$x[, 1:2]) %>%
  rownames_to_column("sample") %>%
  mutate(condition = ifelse(grepl("^BC", sample), "Control", "Experimental"))

pct_var  <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

ggplot(pca_df, aes(x = PC1, y = PC2, colour = condition, label = sample)) +
  geom_point(size = 5, alpha = 0.9) +
  geom_text_repel(size = 4, fontface = "bold") +
  scale_colour_manual(values = c(Control = "#2166ac", Experimental = "#d6604d")) +
  labs(title = "PCA — Colwellia 34H proteome",
       x = paste0("PC1 (", pct_var[1], "%)"),
       y = paste0("PC2 (", pct_var[2], "%)"),
       colour = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

## 7.2 Sample–sample correlation heatmap

```{r cor-heatmap, fig.cap="Fig 3. Pearson correlation of median-centred log2(NSAF) between all samples."}
cor_mat <- cor(log2_norm, method = "pearson")

ann_col <- data.frame(
  Condition = ifelse(grepl("^BC", colnames(cor_mat)), "Control", "Experimental"),
  row.names = colnames(cor_mat)
)
ann_colors <- list(Condition = c(Control = "#2166ac", Experimental = "#d6604d"))

pheatmap(cor_mat,
         annotation_col  = ann_col,
         annotation_colors = ann_colors,
         color           = colorRampPalette(c("#f7fbff","#2171b5"))(100),
         display_numbers = TRUE,
         number_format   = "%.3f",
         fontsize_number = 9,
         main            = "Sample–sample Pearson correlation",
         border_color    = NA)
```

## 7.3 Volcano plot

```{r volcano, fig.height=7, fig.cap="Fig 4. Volcano plot. Red = FDR < 0.05, grey = not significant. Top 15 hits labelled."}
sig_cutoff <- 0.05

vol_df <- results_tbl %>%
  mutate(
    sig       = adj_p_BH < sig_cutoff,
    direction = case_when(
      sig & logFC  > 0 ~ "Up in Exp",
      sig & logFC  < 0 ~ "Down in Exp",
      TRUE             ~ "NS"
    ),
    # Short label: first description before comma or bracket
    short_desc = str_extract(`Protein Description(s)`, "^[^,\\[]+") %>%
                 str_trunc(45)
  )

top_labels <- vol_df %>%
  filter(sig) %>%
  slice_min(adj_p_BH, n = 15)

ggplot(vol_df, aes(x = logFC, y = -log10(p_value), colour = direction)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_text_repel(data = top_labels,
                  aes(label = short_desc),
                  size = 2.8, max.overlaps = 20,
                  segment.colour = "grey60") +
  geom_hline(yintercept = -log10(max(results_tbl$p_value[results_tbl$adj_p_BH < sig_cutoff])),
             linetype = "dashed", colour = "grey40", linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "solid", colour = "grey70", linewidth = 0.4) +
  scale_colour_manual(values = c("Up in Exp"   = "#d6604d",
                                 "Down in Exp" = "#2166ac",
                                 "NS"          = "grey75")) +
  labs(title  = "Volcano plot — Exp vs Ctrl",
       x      = "log2 Fold Change (Exp − Ctrl)",
       y      = "−log10(p-value)",
       colour = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

## 7.4 MA plot

```{r ma-plot, fig.cap="Fig 5. MA plot — mean log2(NSAF) vs fold change. Red = FDR < 0.05."}
ggplot(results_tbl, aes(x = mean_expr, y = logFC,
                         colour = adj_p_BH < sig_cutoff)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "#d6604d"),
                      labels = c("NS", "FDR < 0.05")) +
  labs(title  = "MA plot",
       x      = "Mean log2(NSAF)",
       y      = "log2 Fold Change (Exp − Ctrl)",
       colour = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

## 7.5 Heatmap of significant proteins

```{r heatmap-sig, fig.height=10, fig.cap="Fig 6. Heatmap of FDR < 0.05 proteins. Z-scored across samples."}
sig_ids <- results_tbl %>%
  filter(adj_p_BH < sig_cutoff) %>%
  pull(protein_id)

heat_mat <- log2_norm[sig_ids, ]

# Z-score each row for visualisation
heat_z <- t(scale(t(heat_mat)))

# Short row labels
row_labels <- results_tbl %>%
  filter(protein_id %in% sig_ids) %>%
  mutate(lbl = str_extract(`Protein Description(s)`, "^[^,\\[]+") %>%
               str_trunc(50)) %>%
  pull(lbl)
names(row_labels) <- sig_ids

ann_col2 <- data.frame(
  Condition = c("Control","Control","Control","Experimental","Experimental","Experimental"),
  row.names = colnames(heat_z)
)

pheatmap(heat_z,
         annotation_col   = ann_col2,
         annotation_colors = ann_colors,
         show_rownames    = TRUE,
         labels_row       = row_labels[rownames(heat_z)],
         fontsize_row     = 6,
         fontsize_col     = 10,
         color            = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         main             = paste0("Significant proteins (FDR < 0.05, n = ",
                                   length(sig_ids), ") — Z-score"),
         border_color     = NA)
```

## 7.6 Top hits barplot

```{r barplot-tophits, fig.height=7, fig.cap="Fig 7. Top 20 most significant proteins by adjusted p-value, coloured by direction."}
top20 <- results_tbl %>%
  slice_min(adj_p_BH, n = 20) %>%
  mutate(
    short_desc = str_extract(`Protein Description(s)`, "^[^,\\[]+") %>%
                 str_trunc(55),
    short_desc = fct_reorder(short_desc, logFC)
  )

ggplot(top20, aes(x = logFC, y = short_desc,
                   fill = logFC > 0)) +
  geom_col(alpha = 0.85) +
  geom_vline(xintercept = 0, colour = "grey40") +
  scale_fill_manual(values = c("TRUE" = "#d6604d", "FALSE" = "#2166ac"),
                    labels = c("Down in Exp", "Up in Exp")) +
  labs(title  = "Top 20 differentially abundant proteins",
       x      = "log2 Fold Change (Exp − Ctrl)",
       y      = NULL,
       fill   = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top",
        axis.text.y     = element_text(size = 8))
```

---

# 8 · Export results

```{r export}
# Full results table sorted by adjusted p-value
results_export <- results_tbl %>%
  arrange(adj_p_BH) %>%
  select(`Protein(s)`, `Protein Description(s)`,
         logFC, mean_expr, mod_t, p_value, adj_p_BH, B_statistic)

# Add individual normalised NSAF columns
nsaf_export <- as.data.frame(log2_norm) %>%
  rownames_to_column("protein_id")
results_export <- left_join(
  results_export %>% mutate(protein_id = results_tbl$protein_id[
    match(results_tbl$adj_p_BH, results_export$adj_p_BH)]),   # re-attach id
  nsaf_export, by = "protein_id") %>%
  select(-protein_id)

# Write to Excel
wb <- createWorkbook()
addWorksheet(wb, "limma_results")
writeData(wb, "limma_results", results_export)

addWorksheet(wb, "summary")
writeData(wb, "summary", data.frame(
  Metric = c("Proteins tested", "FDR<0.05 (total)",
             "FDR<0.05 up in Exp", "FDR<0.05 down in Exp",
             "FDR<0.01 (total)",
             "Prior df (d0)", "Prior variance (s0²)",
             "Posterior df"),
  Value  = c(nrow(results_tbl),
             sum(results_tbl$adj_p_BH < 0.05),
             sum(results_tbl$adj_p_BH < 0.05 & results_tbl$logFC > 0),
             sum(results_tbl$adj_p_BH < 0.05 & results_tbl$logFC < 0),
             sum(results_tbl$adj_p_BH < 0.01),
             round(fit_eb$df.prior, 2),
             round(fit_eb$s2.prior, 6),
             round(fit_eb$df.prior + fit_eb$df.residual[1], 2))
))

saveWorkbook(wb, "Bacteria_limma_results_R.xlsx", overwrite = TRUE)
cat("Results saved to Bacteria_limma_results_R.xlsx\n")
```

---

# 9 · Session info

```{r session-info}
sessionInfo()
```
