---
title: "DNA Shield Proteomics — Peptide Filtering"
author: "Evelyn Bartlett"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
```

## Overview

This document describes the peptide filtering pipeline applied to two Limelight/Comet-derived peptide lists from the DNA Shield proteomics experiment:

- **Bacteria** (*Colwellia* 34H) — `peptides (7).txt`
- **Phytoplankton** (*Thalassiosira pseudonana*) — `peptides (8).txt`

Two filtering steps were applied:

1. **Contaminant removal** — peptides mapping exclusively to common laboratory contaminants (cRAP database: trypsin, keratins, albumin, etc.)
2. **Cross-contamination removal** — peptides whose bare sequence (modifications stripped) appeared in both datasets

---

## Load Filtered Data

```{r load}
bacteria   <- read_csv("bacteria_peptides_filtered.csv")
phyto      <- read_csv("phytoplankton_peptides_filtered.csv")

cat("Bacteria peptides:      ", nrow(bacteria), "\n")
cat("Phytoplankton peptides: ", nrow(phyto), "\n")
```

---

## Filtering Summary

| Step | Bacteria removed | Phytoplankton removed |
|------|-----------------|----------------------|
| Raw input | — (33,221) | — (32,835) |
| Contaminant removal | 139 | 79 |
| Cross-contamination removal | 2,117 | 2,117 |
| **Final** | **30,711** | **30,121** |

### Contaminant patterns targeted

Common cRAP contaminants flagged for removal:

```{r contam_list, echo=FALSE}
contaminants <- c(
  "TRYP_PIG / P00761 (Porcine trypsin)",
  "KRT* / K2C* / K1C* (Keratins)",
  "ALBU_ / P02769 (Serum albumin)",
  "CTRA_ / CTRB_ (Chymotrypsin)",
  "LYSC_ (Lysozyme)",
  "RNAS1_ (RNase A)",
  "TRFE_ (Transferrin)",
  "CON__ (MaxQuant contaminant prefix)"
)
knitr::kable(data.frame(Contaminant = contaminants), col.names = "Contaminant patterns removed")
```

---

## PSM Count Distributions

```{r psm_bact, fig.cap="CP34H: distribution of total PSM counts per peptide"}
bact_psm_cols <- grep("PSM Count", names(bacteria), value = TRUE)

bacteria %>%
  mutate(total_PSM = rowSums(across(all_of(bact_psm_cols)), na.rm = TRUE)) %>%
  ggplot(aes(x = total_PSM)) +
  geom_histogram(bins = 80, fill = "#2166ac", color = "white", alpha = 0.85) +
  scale_x_log10() +
  labs(
    title = "CP34H (Colwellia 34H) — Total PSM per Peptide",
    x = "Total PSM (log10 scale)",
    y = "Number of peptides"
  ) +
  theme_minimal(base_size = 13)
```

```{r psm_phyto, fig.cap="Phytoplankton: distribution of total PSM counts per peptide"}
phyto_psm_cols <- grep("PSM Count", names(phyto), value = TRUE)

phyto %>%
  mutate(total_PSM = rowSums(across(all_of(phyto_psm_cols)), na.rm = TRUE)) %>%
  ggplot(aes(x = total_PSM)) +
  geom_histogram(bins = 80, fill = "#1b7837", color = "white", alpha = 0.85) +
  scale_x_log10() +
  labs(
    title = "Phytoplankton (T. pseudonana) — Total PSM per Peptide",
    x = "Total PSM (log10 scale)",
    y = "Number of peptides"
  ) +
  theme_minimal(base_size = 13)
```

---

## Unique vs. Non-Unique Peptides

```{r unique}
bact_unique  <- bacteria %>% count(Unique) %>% mutate(organism = "Bacteria")
phyto_unique <- phyto    %>% count(Unique) %>% mutate(organism = "Phytoplankton")

bind_rows(bact_unique, phyto_unique) %>%
  mutate(Unique = ifelse(Unique == "*", "Unique", "Non-unique")) %>%
  ggplot(aes(x = organism, y = n, fill = Unique)) +
  geom_col(position = "dodge", alpha = 0.9) +
  scale_fill_manual(values = c("Unique" = "#4d9de0", "Non-unique" = "#aab4be")) +
  labs(
    title = "Unique vs. Non-unique Peptides After Filtering",
    x = NULL, y = "Number of peptides", fill = NULL
  ) +
  theme_minimal(base_size = 13)
```

---

## Discussion

Peptide lists were exported from Limelight following Comet database searches with PSM- and peptide-level FDR ≤ 0.01. No protein-level FDR filter was applied upstream.

**Contaminant removal** targeted 139 bacteria-assigned and 79 phytoplankton-assigned peptide entries matching cRAP database identifiers. The dominant contaminants were porcine trypsin (TRYP_PIG) and human/bovine keratins, consistent with typical bottom-up proteomics sample preparation.

**Cross-contamination removal** identified 2,117 peptide sequences (bare, modification-agnostic) present in both datasets. Given that these experiments co-culture *Colwellia* 34H with *T. pseudonana*, shared peptides could reflect true biological sequence conservation or genuine cross-species carry-over. To be conservative, all shared sequences were excluded from both lists to ensure organism-specific downstream analyses.

The final filtered lists (30,711 bacteria; 30,121 phytoplankton) are saved as:

- `bacteria_peptides_filtered.csv`
- `phytoplankton_peptides_filtered.csv`

---

## Session Info

```{r session}
sessionInfo()
```
