---
title: "Condition-Unique Peptide Landscape — Nature Figure"
subtitle: "Colwellia 34H (BC vs E) | T. pseudonana (C vs D1:2)"
author: "Evelyn Bartlett — Brook Lab"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: false
    theme: minimal
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi     = 300,
  fig.width  = 7.2,   # Nature double-column = 183 mm ≈ 7.2 in
  fig.height = 8.5
)

# ── Packages ────────────────────────────────────────────────────────────────
pkgs <- c("tidyverse", "ggridges", "patchwork", "ggbeeswarm")
invisible(lapply(pkgs, library, character.only = TRUE))

# ── Nature-style theme ──────────────────────────────────────────────────────
theme_nature <- function(base_size = 7) {
  theme_classic(base_size = base_size) +
    theme(
      text             = element_text(family = "Helvetica", colour = "black"),
      axis.line        = element_line(colour = "black", linewidth = 0.4),
      axis.ticks       = element_line(colour = "black", linewidth = 0.4),
      axis.ticks.length = unit(2, "pt"),
      axis.text        = element_text(colour = "black", size = base_size),
      axis.title       = element_text(colour = "black", size = base_size + 0.5,
                                      face = "plain"),
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold", size = base_size + 0.5),
      panel.spacing    = unit(4, "pt"),
      legend.key.size  = unit(7, "pt"),
      legend.text      = element_text(size = base_size),
      legend.title     = element_text(size = base_size, face = "bold"),
      plot.title       = element_text(size = base_size + 1, face = "bold",
                                      margin = margin(b = 3)),
      plot.tag         = element_text(size = base_size + 2, face = "bold")
    )
}

# ── Colour palette (poster scheme: bacteria = blue, phyto = green) ─────────
pal <- c(
  BC_only    = "#2166AC",   # dark blue  — bacteria control-unique
  E_only     = "#92C5DE",   # light blue — bacteria experimental-unique
  C_only     = "#1B7837",   # dark green — phyto control-unique
  `D1:2_only` = "#74C476"   # light green — phyto D1:2-unique
)

label_map <- c(
  BC_only     = "BC only\n(bacteria ctrl)",
  E_only      = "E only\n(bacteria exp)",
  C_only      = "C only\n(phyto ctrl)",
  `D1:2_only` = "D1:2 only\n(phyto exp)"
)
```

```{r load-data}
strip_mod <- function(x) str_remove_all(x, "\\[.*?\\]")

missed_cleavage <- function(seq) {
  str_count(seq, "[KR](?!$)")
}

gravy_score <- function(seq) {
  kd <- c(A=1.8,R=-4.5,N=-3.5,D=-3.5,C=2.5,Q=-3.5,E=-3.5,G=-0.4,H=-3.2,
          I=4.5,L=3.8,K=-3.9,M=1.9,F=2.8,P=-1.6,S=-0.8,T=-0.7,W=-0.9,
          Y=-1.3,V=4.2)
  aas <- str_split(seq, "")[[1]]
  mean(kd[aas], na.rm = TRUE)
}

bact  <- read_csv("bacteria_venn_peptides.csv",  show_col_types = FALSE)
phyto <- read_csv("phytoplankton_venn_peptides.csv", show_col_types = FALSE)

bc_raw  <- na.omit(bact$BC)
e_raw   <- na.omit(bact$E)
c_raw   <- na.omit(phyto$C)
d12_raw <- na.omit(phyto$`D1:2`)

make_feat <- function(peps, grp) {
  tibble(peptide = peps, group = grp) %>%
    mutate(
      seq    = strip_mod(peptide),
      length = nchar(seq),
      has_oxM = str_detect(peptide, "M\\[15\\.99\\]"),
      mc     = missed_cleavage(seq),
      gravy  = map_dbl(seq, gravy_score),
      cterm  = str_sub(seq, -1)
    )
}

feat <- bind_rows(
  make_feat(setdiff(bc_raw,  e_raw),   "BC_only"),
  make_feat(setdiff(e_raw,   bc_raw),  "E_only"),
  make_feat(setdiff(c_raw,  d12_raw),  "C_only"),
  make_feat(setdiff(d12_raw, c_raw),   "D1:2_only")
) %>%
  mutate(
    group      = factor(group, levels = names(pal)),
    organism   = if_else(str_detect(group, "^BC|^E"),
                         "Colwellia 34H", "T. pseudonana"),
    condition  = if_else(str_detect(group, "BC|C_"),
                         "Control-unique", "Experimental-unique"),
    mc_cat     = factor(pmin(mc, 2), labels = c("0","1","≥2"))
  )
```

```{r panel-A-aa-composition, fig.height=3}
# ── AA composition log2(control/experimental) lollipop ────────────────────
aa_freq <- feat %>%
  group_by(organism, condition) %>%
  summarise(
    total = sum(nchar(seq)),
    across(all_of(strsplit("ACDEFGHIKLMNPQRSTVWY","")[[1]]),
           ~ 100 * sum(str_count(seq, .x)) / total),
    .groups = "drop"
  ) %>%
  pivot_longer(A:Y, names_to = "aa", values_to = "pct")

aa_diff <- aa_freq %>%
  pivot_wider(names_from = condition, values_from = pct) %>%
  mutate(
    log2fc    = log2((`Control-unique` + 0.01) / (`Experimental-unique` + 0.01)),
    highlight = case_when(
      aa == "C" & organism == "T. pseudonana" ~ "Cys (redox)",
      aa == "H" & organism == "Colwellia 34H" ~ "His (metal-binding)",
      aa %in% c("R","G") & organism == "Colwellia 34H" ~ "Arg/Gly (ribosomal)",
      TRUE ~ "other"
    ),
    aa = fct_reorder(aa, log2fc)
  )

cols_hi <- c(
  "Cys (redox)"         = "#E31A1C",
  "His (metal-binding)" = "#FF7F00",
  "Arg/Gly (ribosomal)" = "#6A3D9A",
  "other"               = "grey70"
)

pA <- ggplot(aa_diff, aes(log2fc, aa)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3,
             colour = "grey50") +
  geom_segment(aes(x = 0, xend = log2fc, y = aa, yend = aa,
                   colour = highlight), linewidth = 0.55) +
  geom_point(aes(colour = highlight, size = highlight == "other"),
             show.legend = TRUE) +
  scale_size_manual(values = c(`TRUE` = 1.2, `FALSE` = 2.0), guide = "none") +
  scale_colour_manual(values = cols_hi, name = NULL,
                      breaks = c("Cys (redox)","His (metal-binding)","Arg/Gly (ribosomal)")) +
  facet_wrap(~organism, scales = "free_x") +
  labs(
    x     = expression(log[2]~(control-unique / experimental-unique)),
    y     = "Amino acid",
    title = "a  Amino acid composition bias in condition-unique peptides"
  ) +
  theme_nature() +
  theme(legend.position = "right")

pA
```

```{r panel-B-length-ridge, fig.height=2.8}
# ── Peptide length ridge plot ──────────────────────────────────────────────
pB <- ggplot(feat, aes(x = length, y = group, fill = group)) +
  geom_density_ridges(
    aes(height = after_stat(density)),
    stat = "density", trim = TRUE,
    alpha = 0.82, scale = 1.1,
    linewidth = 0.3, colour = "white",
    rel_min_height = 0.005
  ) +
  geom_vline(
    data = feat %>% group_by(group) %>% summarise(med = median(length)),
    aes(xintercept = med, colour = group),
    linetype = "dashed", linewidth = 0.45, show.legend = FALSE
  ) +
  scale_fill_manual(values = pal, guide = "none") +
  scale_colour_manual(values = pal, guide = "none") +
  scale_x_continuous(limits = c(5, 45), breaks = seq(10,40,10)) +
  scale_y_discrete(labels = label_map) +
  labs(
    x     = "Peptide length (amino acids)",
    y     = NULL,
    title = "b  Peptide length distributions"
  ) +
  theme_nature() +
  theme(axis.text.y = element_text(size = 6))

pB
```

```{r panel-C-properties, fig.height=3}
# ── Dumbbell: key peptide properties ──────────────────────────────────────
prop_sum <- feat %>%
  group_by(organism, condition, group) %>%
  summarise(
    mean_length  = mean(length),
    pct_mc       = 100 * mean(mc >= 1),
    pct_oxM      = 100 * mean(has_oxM),
    mean_gravy   = mean(gravy),
    .groups = "drop"
  ) %>%
  pivot_longer(mean_length:mean_gravy, names_to = "metric", values_to = "value") %>%
  mutate(
    metric = recode(metric,
      mean_length = "Mean peptide\nlength (aa)",
      pct_mc      = "Peptides with\n≥1 missed cleavage (%)",
      pct_oxM     = "Met-oxidised\npeptides (%)",
      mean_gravy  = "Mean GRAVY\nscore"
    ),
    metric = factor(metric, levels = c(
      "Mean peptide\nlength (aa)",
      "Peptides with\n≥1 missed cleavage (%)",
      "Met-oxidised\npeptides (%)",
      "Mean GRAVY\nscore"
    ))
  )

# Dumbbell: one line per organism per metric, points = ctrl vs exp
dumbbell_df <- prop_sum %>%
  select(organism, condition, metric, value) %>%
  pivot_wider(names_from = condition, values_from = value)

pC <- ggplot(prop_sum, aes(x = value, y = metric)) +
  # connector line
  geom_line(data = dumbbell_df,
            aes(x = `Control-unique`, xend = `Experimental-unique`,
                y = metric, group = metric),
            inherit.aes = FALSE,
            colour = "grey60", linewidth = 0.5) +
  geom_line(data = dumbbell_df,
            aes(x = `Experimental-unique`, y = metric, group = metric),
            inherit.aes = FALSE,
            colour = "grey60", linewidth = 0.5) +
  # use a segment geom for the dumbbell connector
  geom_segment(
    data = dumbbell_df,
    aes(x = `Control-unique`, xend = `Experimental-unique`,
        y = metric, yend = metric),
    inherit.aes = FALSE,
    colour = "grey70", linewidth = 0.6
  ) +
  geom_point(aes(colour = group), size = 2.5) +
  scale_colour_manual(values = pal, labels = label_map, name = NULL) +
  facet_wrap(~organism, scales = "free_x") +
  labs(
    x     = "Value",
    y     = NULL,
    title = "c  Biochemical properties of condition-unique peptides"
  ) +
  theme_nature() +
  theme(
    legend.position = "bottom",
    legend.key.size = unit(8, "pt"),
    axis.text.y     = element_text(size = 6)
  ) +
  guides(colour = guide_legend(nrow = 2))

pC
```

```{r assemble-figure, fig.width=7.2, fig.height=9}
# ── Final assembled figure ─────────────────────────────────────────────────
final_fig <- pA / pB / pC +
  plot_layout(heights = c(3, 2.5, 3)) &
  theme(plot.margin = margin(4, 6, 4, 6))

final_fig
```

```{r save-figure}
# Save as 300 dpi TIFF (Nature submission standard) and PDF
ggsave(
  "figure_unique_peptides_nature.tiff",
  plot   = final_fig,
  width  = 7.2, height = 9,
  units  = "in", dpi = 300,
  compression = "lzw"
)

ggsave(
  "figure_unique_peptides_nature.pdf",
  plot   = final_fig,
  width  = 7.2, height = 9,
  units  = "in", device = cairo_pdf
)

cat("Saved: figure_unique_peptides_nature.tiff (.pdf)\n")
```

---

### Figure legend

**Biochemical landscape of condition-unique peptides in *Colwellia* 34H and *T. pseudonana* proteomes.**
**(a)** Log₂ fold-change in per-residue amino acid frequency between control-unique and experimental-unique peptide sets for each organism. Positive values indicate enrichment in control-unique peptides; negative values indicate enrichment in experimental-unique peptides. Highlighted residues denote amino acids with known functional significance: Cys (redox-active proteins; thioredoxins, glutaredoxins), His (metal co-ordination), and Arg/Gly (ribosomal and translation-associated proteins).
**(b)** Kernel density distributions of peptide length for each condition-unique set; dashed lines indicate group medians.
**(c)** Dumbbell plot comparing four biochemical properties — mean peptide length, proportion of peptides with ≥1 trypsin missed cleavage, proportion of Met-oxidised peptides, and mean GRAVY hydrophobicity score — between control-unique (filled circles) and experimental-unique (open circles) sets for each organism. Connecting segments highlight the direction and magnitude of the shift between conditions. All peptides passed a q-value threshold of ≤0.01 at the PSM and peptide level; common laboratory contaminants were excluded prior to analysis.
