---
title: "DNA Shield Proteomics — Identification Counts QC"
subtitle: "Total proteins, peptides, and PSMs per sample type | CP34H & *T. pseudonana*"
author: "Evelyn Bartlett — Brook Lab"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.path   = "figures/",
  dev        = "png",
  dpi        = 600,
  fig.width  = 9,
  fig.height = 4.0
)
```

# Overview

This notebook summarises identification counts (unique proteins, peptides,
and total PSMs) across DNA Shield dilution conditions for both organisms:

| Organism           | Groups                              |
|--------------------|-------------------------------------|
| *Colwellia* CP34H  | B-CTL, B-1.5, B-1.75, B-2.0        |
| *T. pseudonana*    | P-CTL, P-1.5, P-2.0                 |

**n = 3 replicates per group.** Bars show group mean; error bars are ± SEM.
Individual replicate values are overlaid as jittered points. Welch's t-tests
(unequal variance, two-sided) compare each dilution to the within-organism
control; significance is shown as brackets (ns / * / ** / ***).

Contaminants (human keratins, trypsin, bovine albumin, and related common
MS contaminants) are removed before any counting. Peptides present in only
one sample across all replicates (< 2 samples with PSM > 0) are also excluded.

---

# Setup

```{r packages}
library(tidyverse)
library(stringr)
library(ggplot2)
library(patchwork)
library(ggsignif)
library(scales)
```

```{r theme}
# Nature-quality theme — 8 pt sans, no chart-junk
theme_nature <- function(base_size = 8) {
  theme_classic(base_size = base_size, base_family = "sans") +
  theme(
    axis.line        = element_line(linewidth = 0.4),
    axis.ticks       = element_line(linewidth = 0.4),
    axis.text        = element_text(color = "black"),
    axis.text.x      = element_text(angle = 20, hjust = 1),
    plot.title       = element_text(face = "bold", hjust = 0,
                                    size = base_size + 1),
    legend.position  = "none",
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold")
  )
}

save_png <- function(plot, name, w = 9, h = 4.0, dpi = 600,
                     fig_dir = "figures") {
  dir.create(fig_dir, showWarnings = FALSE)
  ggsave(file.path(fig_dir, paste0(name, ".png")),
         plot = plot, width = w, height = h, units = "in",
         dpi = dpi, bg = "white", type = "cairo")
  invisible(plot)
}
```

---

# Contaminant filtering

Common MS contaminants are flagged by regex on the `Protein(s)` accession
string. Any peptide whose *only* proteins are contaminants is dropped.
Peptides that map to both a contaminant and a target protein are kept
(the contaminant accession is removed from the `Protein(s)` string first).

```{r contam_setup}
# Regex patterns applied to the Protein(s) field
contam_re <- paste(
  "K1H1_HUMAN|K1H2_HUMAN|K2C1_HUMAN|K1C9_HUMAN|K1C10_HUMAN",  # keratins
  "K[12][A-Z][0-9]+_HUMAN",   # broad keratin family
  "ALBU_HUMAN|ALBU_BOVIN|serum.albumin|BSA",
  "TRYP_PIG|P00761",           # porcine trypsin
  "cRAP|CON__",
  "KRT[0-9]|KRT[0-9][0-9]",   # keratin gene names sometimes in accessions
  "sp\\|[A-Z0-9]+\\|[A-Z0-9]+_HUMAN",   # any human SwissProt entry
  sep = "|"
)

# Returns TRUE if ALL proteins for a peptide are contaminants
is_contam <- function(prot_str) {
  ids <- str_split(prot_str, ",")[[1]] %>% trimws()
  all(str_detect(ids, contam_re))
}

filter_contaminants <- function(df) {
  n_before <- nrow(df)
  df_clean <- df %>%
    rowwise() %>%
    filter(!is_contam(`Protein(s)`)) %>%
    ungroup()
  n_after <- nrow(df_clean)
  cat(sprintf("  Contaminant filter: removed %d / %d peptides (%.1f%%)\n",
              n_before - n_after, n_before,
              100 * (n_before - n_after) / n_before))
  df_clean
}
```

---

# Helper: compute per-sample identification counts

```{r helpers}
# Count unique protein accessions in rows where PSM count > 0
count_proteins <- function(df, psm_col) {
  df %>%
    filter(.data[[psm_col]] > 0) %>%
    pull(`Protein(s)`) %>%
    str_split(",") %>%
    unlist() %>%
    trimws() %>%
    unique() %>%
    length()
}

# Build a tidy per-sample summary table
#   df        : filtered peptide data frame
#   psm_cols  : named character vector  (name = sample ID, value = column in df)
#   groups    : named list: group label → character vector of sample IDs
build_counts <- function(df, psm_cols, groups) {
  rows <- imap_dfr(groups, function(samples, grp_label) {
    map_dfr(samples, function(s) {
      col <- psm_cols[[s]]
      tibble(
        group      = grp_label,
        sample     = s,
        n_proteins = count_proteins(df, col),
        n_peptides = sum(df[[col]] > 0, na.rm = TRUE),
        total_PSMs = sum(df[[col]],     na.rm = TRUE)
      )
    })
  })
  rows
}

# Build a bar + jitter + signif figure for one metric
#   counts_df    : output of build_counts()
#   metric       : column name (string)
#   y_label      : y-axis label
#   fill_palette : named colour vector (group label → hex)
#   group_order  : factor order for x-axis
#   ctrl_label   : name of the control group (reference for t-tests)
make_panel <- function(counts_df, metric, y_label,
                       fill_palette, group_order, ctrl_label) {

  grp_sum <- counts_df %>%
    group_by(group) %>%
    summarise(
      mean_val = mean(.data[[metric]]),
      sem_val  = sd(.data[[metric]]) / sqrt(n()),
      .groups  = "drop"
    ) %>%
    mutate(group = factor(group, levels = group_order))

  counts_df <- counts_df %>%
    mutate(group = factor(group, levels = group_order))

  # Comparisons: control vs each dilution
  treat_groups <- setdiff(group_order, ctrl_label)
  comparisons  <- lapply(treat_groups, function(g) c(ctrl_label, g))

  ggplot(grp_sum, aes(x = group, y = mean_val, fill = group)) +
    geom_col(color = "black", linewidth = 0.4, width = 0.7) +
    geom_errorbar(aes(ymin = mean_val - sem_val,
                      ymax = mean_val + sem_val),
                  width = 0.18, linewidth = 0.5) +
    geom_jitter(data = counts_df,
                aes(x = group, y = .data[[metric]]),
                width = 0.10, shape = 21, fill = "white",
                color = "black", stroke = 0.4, size = 1.8,
                inherit.aes = FALSE) +
    geom_signif(data        = counts_df,
                aes(x = group, y = .data[[metric]]),
                comparisons = comparisons,
                test        = "t.test",
                test.args   = list(var.equal = FALSE),
                map_signif_level = TRUE,
                step_increase    = 0.10,
                tip_length       = 0.01,
                vjust            = 0.3,
                textsize         = 2.6,
                inherit.aes      = FALSE) +
    scale_fill_manual(values = fill_palette) +
    scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.25))) +
    labs(x = NULL, y = y_label) +
    theme_nature()
}
```

---

# CP34H (*Colwellia psychrerythraea* 34H)

## Load & filter

```{r bact_load}
# bacteria_peptides_filtered_2rep.csv = original filtered file with ≥2-sample filter applied
# Adjust path if running from a different working directory
bact_file <- file.path("..", "..", "..", "..", "bacteria_peptides_filtered_2rep.csv")
if (!file.exists(bact_file))
  bact_file <- "bacteria_peptides_filtered_2rep.csv"

bact_raw <- read_csv(bact_file, show_col_types = FALSE)
cat("Loaded", nrow(bact_raw), "peptides (bacteria)\n")

bact_raw[is.na(bact_raw)] <- 0
cat("Filtering contaminants...\n")
bact <- filter_contaminants(bact_raw)
```

## Configuration

```{r bact_config}
# Map short sample IDs → exact column names in the CSV
bact_psm_cols <- c(
  "BC1" = "PSM Count (HR_BC1)",
  "BC2" = "PSM Count (HR_BC2)",
  "BC3" = "PSM Count (HR_BC3)",
  "E1"  = "PSM Count (HR-E1)",
  "E2"  = "PSM Count (HR_E2)",
  "E3"  = "PSM Count (HR_E3)",
  "F1"  = "PSM Count (HR_F1)",
  "F2"  = "PSM Count (HR-F2)",
  "F3"  = "PSM Count (HR_F3)",
  "G1"  = "PSM Count (HR_G1)",
  "G2"  = "PSM Count (HR_G2)",
  "G3"  = "PSM Count (HR_G3)"
)

# Group membership: display label → sample IDs
bact_groups <- list(
  "B-CTL"  = c("BC1", "BC2", "BC3"),
  "B-1.5"  = c("E1",  "E2",  "E3"),
  "B-1.75" = c("F1",  "F2",  "F3"),
  "B-2.0"  = c("G1",  "G2",  "G3")
)
bact_order <- c("B-CTL", "B-1.5", "B-1.75", "B-2.0")

# Blue family — control is neutral anchor, treatments darken with dilution
bact_palette <- c(
  "B-CTL"  = "#4D4D4D",   # neutral dark grey
  "B-1.5"  = "#AED6F1",   # light blue
  "B-1.75" = "#3498DB",   # medium blue
  "B-2.0"  = "#1B4F72"    # deep blue
)
```

## Compute counts

```{r bact_counts}
bact_counts <- build_counts(bact, bact_psm_cols, bact_groups)
print(bact_counts)
```

## Welch t-test summary table

```{r bact_ttest}
run_welch <- function(counts_df, metric, ctrl_label, group_order) {
  ctrl_vals <- counts_df %>% filter(group == ctrl_label) %>% pull(!!sym(metric))
  treat_groups <- setdiff(group_order, ctrl_label)

  map_dfr(treat_groups, function(g) {
    treat_vals <- counts_df %>% filter(group == g) %>% pull(!!sym(metric))
    res <- t.test(treat_vals, ctrl_vals, var.equal = FALSE)
    tibble(
      comparison = paste0(g, " vs ", ctrl_label),
      metric     = metric,
      mean_ctrl  = mean(ctrl_vals),
      mean_treat = mean(treat_vals),
      delta      = mean(treat_vals) - mean(ctrl_vals),
      pct_change = 100 * (mean(treat_vals) - mean(ctrl_vals)) / mean(ctrl_vals),
      t_stat     = unname(res$statistic),
      df         = unname(res$parameter),
      p_value    = res$p.value,
      sig        = case_when(
        res$p.value < 0.001 ~ "***",
        res$p.value < 0.01  ~ "**",
        res$p.value < 0.05  ~ "*",
        TRUE                ~ "ns"
      )
    )
  })
}

bact_tests <- bind_rows(
  run_welch(bact_counts, "n_proteins", "B-CTL", bact_order),
  run_welch(bact_counts, "n_peptides", "B-CTL", bact_order),
  run_welch(bact_counts, "total_PSMs", "B-CTL", bact_order)
)
print(bact_tests)
write_csv(bact_tests, "bact_identification_welch_tests.csv")
```

## Figure 1 — CP34H identification counts

```{r fig_bact, fig.width=9, fig.height=4.0}
p_bact_prot <- make_panel(bact_counts, "n_proteins",
                           "Proteins identified\n(mean ± SEM)",
                           bact_palette, bact_order, "B-CTL") +
  labs(title = "a  Unique proteins")

p_bact_pep  <- make_panel(bact_counts, "n_peptides",
                           "Peptides identified\n(mean ± SEM)",
                           bact_palette, bact_order, "B-CTL") +
  labs(title = "b  Unique peptides")

p_bact_psm  <- make_panel(bact_counts, "total_PSMs",
                           "Total PSMs\n(mean ± SEM)",
                           bact_palette, bact_order, "B-CTL") +
  labs(title = "c  Total PSMs")

fig_bact <- p_bact_prot + p_bact_pep + p_bact_psm +
  plot_annotation(
    title    = expression(italic("Colwellia psychrerythraea") ~ "34H — Identification counts by condition"),
    subtitle = "n = 3 replicates per group | bars = mean ± SEM | Welch's t-test vs B-CTL",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 9),
      plot.subtitle = element_text(size = 7, color = "grey40")
    )
  )

fig_bact
save_png(fig_bact, "fig_CP34H_identification_counts", w = 9, h = 4.0)
```

---

# *Thalassiosira pseudonana*

## Load & filter

```{r phyto_load}
# phytoplankton_peptides_filtered_PED6_2rep.csv = re-filtered from raw Limelight export,
# including PED6 (1:2 replicate 3), contaminants removed, Colwellia cross-contams removed,
# and ≥2-sample filter applied.
phyto_file <- file.path("..", "..", "..", "..", "phytoplankton_peptides_filtered_PED6_2rep.csv")
if (!file.exists(phyto_file))
  phyto_file <- "phytoplankton_peptides_filtered_PED6_2rep.csv"

phyto_raw <- read_csv(phyto_file, show_col_types = FALSE)
cat("Loaded", nrow(phyto_raw), "peptides (phytoplankton)\n")

phyto_raw[is.na(phyto_raw)] <- 0
cat("Filtering contaminants...\n")
phyto <- filter_contaminants(phyto_raw)
```

## Configuration

```{r phyto_config}
phyto_psm_cols <- c(
  "PEC1"  = "PSM Count (PEC1)",
  "PEC2"  = "PSM Count (PEC2)",
  "PEC3"  = "PSM Count (PEC3)",
  "PED1"  = "PSM Count (PED1 1.5)",
  "PED2"  = "PSM Count (PED2 1.5)",
  "PED3"  = "PSM Count (PED3 1.5)",
  "PED4"  = "PSM Count (PED4 1:2)",
  "PED5"  = "PSM Count (PED5 1:2)",
  "PED6"  = "PSM Count (PED6 1:2)"
)

phyto_groups <- list(
  "P-CTL" = c("PEC1", "PEC2", "PEC3"),
  "P-1.5" = c("PED1", "PED2", "PED3"),
  "P-2.0" = c("PED4", "PED5", "PED6")   # n = 3
)
phyto_order <- c("P-CTL", "P-1.5", "P-2.0")

# Green family — control is neutral anchor, treatments darken with dilution
phyto_palette <- c(
  "P-CTL" = "#4D4D4D",   # neutral dark grey
  "P-1.5" = "#A9DFBF",   # light green
  "P-2.0" = "#1E8449"    # deep green
)
```

## Compute counts

```{r phyto_counts}
phyto_counts <- build_counts(phyto, phyto_psm_cols, phyto_groups)
print(phyto_counts)
```

## Welch t-test summary table

```{r phyto_ttest}
phyto_tests <- bind_rows(
  run_welch(phyto_counts, "n_proteins", "P-CTL", phyto_order),
  run_welch(phyto_counts, "n_peptides", "P-CTL", phyto_order),
  run_welch(phyto_counts, "total_PSMs", "P-CTL", phyto_order)
)
print(phyto_tests)
write_csv(phyto_tests, "phyto_identification_welch_tests.csv")
```

## Figure 2 — *T. pseudonana* identification counts

```{r fig_phyto, fig.width=7.5, fig.height=4.0}
p_phyto_prot <- make_panel(phyto_counts, "n_proteins",
                            "Proteins identified\n(mean ± SEM)",
                            phyto_palette, phyto_order, "P-CTL") +
  labs(title = "a  Unique proteins")

p_phyto_pep  <- make_panel(phyto_counts, "n_peptides",
                            "Peptides identified\n(mean ± SEM)",
                            phyto_palette, phyto_order, "P-CTL") +
  labs(title = "b  Unique peptides")

p_phyto_psm  <- make_panel(phyto_counts, "total_PSMs",
                            "Total PSMs\n(mean ± SEM)",
                            phyto_palette, phyto_order, "P-CTL") +
  labs(title = "c  Total PSMs")

fig_phyto <- p_phyto_prot + p_phyto_pep + p_phyto_psm +
  plot_annotation(
    title    = expression(italic("Thalassiosira pseudonana") ~ "— Identification counts by condition"),
    subtitle = "n = 3 replicates per group | bars = mean ± SEM | Welch's t-test vs P-CTL",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 9),
      plot.subtitle = element_text(size = 7, color = "grey40")
    )
  )

fig_phyto
save_png(fig_phyto, "fig_TPs_identification_counts", w = 7.5, h = 4.0)
```

---

# Session info

```{r session_info}
sessionInfo()
```
