---
title: "DNA Shield Proteomics — NSAF Replicate Reproducibility (Figs 5 & 6)"
subtitle: "Colwellia psychrerythraea 34H & Thalassiosira pseudonana | Pairwise correlations · NSAF scatter"
author: "Evelyn Bartlett — Brook Lab"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 8,
  dpi = 300
)
```

# Overview

This script generates **Figures 5 and 6** (replicate reproducibility) in two parts:

- **Part 1 — Pairwise Correlation Plots (panels A & B):** GGally pairwise scatter
  matrix with Pearson *r* labels, computed from protein-level log₂ PSM counts
  aggregated from the peptide-filtered tables.
- **Part 2 — NSAF Replicate Scatter (panel C):** Pairwise log₁₀ NSAF scatter
  with R² annotations, computed from the per-protein NSAF tables
  (`Bacteria/proteins_clean.csv`, `Phytoplankton/proteins_clean.csv`).

Colour scheme follows the Brook Lab poster convention:
**bacteria = blue family** (#2c7fb8 / #1f78b4), **phytoplankton = green family** (#2ca25f / #1A7E3B).

Contaminants (keratins, trypsin, albumin, human/bovine sequences, cRAP tags)
are removed before all analyses.

Upstream pipeline: Comet 2023.02 via Limelight, PSM- and peptide-level q ≤ 0.01,
no protein-level FDR applied.

---

```{r packages}
library(tidyverse)
library(ggplot2)
library(GGally)      # Part 1
library(patchwork)   # Part 2
library(scales)
```

---

# Part 1 — Pairwise Correlation Plots (Figs 5A–B, 6A–B)

Protein-level log₂ PSM counts from peptide-filtered tables. Each peptide is
assigned to its first-listed protein; PSMs are summed per protein per sample.

## Shared helpers

```{r p1_helpers}
# Contaminant filter (applied to Protein(s) accession column)
contam_pattern <- regex(
  "keratin|trypsin|albumin|bsa|contam|KRT|fibrinogen|myosin|actin|
   immunoglobulin|serum|casein|collagen|elastin",
  ignore_case = TRUE
)

# Pearson r label for upper panels
upper_pearson <- function(colour) {
  function(data, mapping, ...) {
    x_col <- as.character(mapping$x); y_col <- as.character(mapping$y)
    x <- data[[x_col]]; y <- data[[y_col]]
    keep <- is.finite(x) & is.finite(y) & (x > 0 | y > 0)
    r <- cor(x[keep], y[keep], method = "pearson")
    ggplot(data = data[keep, ], mapping = mapping) +
      geom_blank() +
      annotate("text", x = mean(range(x[keep])), y = mean(range(y[keep])),
               label = sprintf("r = %.4f", r), size = 5, fontface = "bold",
               colour = colour) +
      theme_void()
  }
}

# Scatter + 1:1 diagonal for lower panels
lower_scatter <- function(colour) {
  function(data, mapping, ...) {
    x_col <- as.character(mapping$x); y_col <- as.character(mapping$y)
    x <- data[[x_col]]; y <- data[[y_col]]
    keep <- is.finite(x) & is.finite(y) & (x > 0 | y > 0)
    ggplot(data = data[keep, ], mapping = mapping) +
      geom_point(alpha = 0.35, size = 0.8, colour = colour) +
      geom_abline(intercept = 0, slope = 1,
                  colour = "firebrick", linewidth = 0.6, linetype = "dashed") +
      theme_bw(base_size = 10)
  }
}
```

## 1.1 CP34H (*Colwellia psychrerythraea* 34H)

```{r p1_bact_load}
bact_raw <- read_csv("bacteria_peptides_filtered.csv", show_col_types = FALSE)
bact_clean <- bact_raw %>% filter(!str_detect(`Protein(s)`, contam_pattern))
cat("CP34H peptides — raw:", nrow(bact_raw), "| after contaminant filter:", nrow(bact_clean), "\n")

# Aggregate to protein level (sum PSMs)
bact_proteins <- bact_clean %>%
  mutate(Protein = str_extract(`Protein(s)`, "[^,]+") %>% str_trim()) %>%
  select(Protein, starts_with("PSM Count")) %>%
  group_by(Protein) %>%
  summarise(across(everything(), sum), .groups = "drop") %>%
  rename_with(~ str_remove_all(.x, "PSM Count \\(|\\)"), starts_with("PSM"))

# Log2 transform; remove all-zero proteins
bact_log <- bact_proteins %>%
  mutate(across(-Protein, ~ log2(.x + 1))) %>%
  filter(rowSums(select(., -Protein)) > 0)

cat("Protein groups retained:", nrow(bact_log), "\n")
```

### Condition BC — control triplicates

```{r p1_bact_BC, fig.cap="CP34H — Condition BC pairwise correlations"}
bact_log %>%
  select(all_of(c("HR_BC1", "HR_BC2", "HR_BC3"))) %>%
  ggpairs(
    upper = list(continuous = upper_pearson("#2c7fb8")),
    lower = list(continuous = lower_scatter("#2c7fb8")),
    diag  = list(continuous = wrap("densityDiag", fill = "#a6cee3", colour = "#1f78b4", alpha = 0.7)),
    columnLabels = c("HR_BC1", "HR_BC2", "HR_BC3")
  ) +
  labs(title = "CP34H — Condition BC (log₂ PSM, protein level)") +
  theme_bw(base_size = 11)
```

### Condition E — DNA Shield 1:1.5 triplicates

```{r p1_bact_E, fig.cap="CP34H — Condition E pairwise correlations"}
bact_log %>%
  select(all_of(c("HR-E1", "HR_E2", "HR_E3"))) %>%
  ggpairs(
    upper = list(continuous = upper_pearson("#2c7fb8")),
    lower = list(continuous = lower_scatter("#2c7fb8")),
    diag  = list(continuous = wrap("densityDiag", fill = "#a6cee3", colour = "#1f78b4", alpha = 0.7)),
    columnLabels = c("HR-E1", "HR_E2", "HR_E3")
  ) +
  labs(title = "CP34H — Condition E (log₂ PSM, protein level)") +
  theme_bw(base_size = 11)
```

### Condition F — DNA Shield 1:1.75 triplicates

```{r p1_bact_F, fig.cap="CP34H — Condition F pairwise correlations"}
bact_log %>%
  select(all_of(c("HR_F1", "HR-F2", "HR_F3"))) %>%
  ggpairs(
    upper = list(continuous = upper_pearson("#2c7fb8")),
    lower = list(continuous = lower_scatter("#2c7fb8")),
    diag  = list(continuous = wrap("densityDiag", fill = "#a6cee3", colour = "#1f78b4", alpha = 0.7)),
    columnLabels = c("HR_F1", "HR-F2", "HR_F3")
  ) +
  labs(title = "CP34H — Condition F (log₂ PSM, protein level)") +
  theme_bw(base_size = 11)
```

### Condition G — DNA Shield 1:2 triplicates

```{r p1_bact_G, fig.cap="CP34H — Condition G pairwise correlations"}
bact_log %>%
  select(all_of(c("HR_G1", "HR_G2", "HR_G3"))) %>%
  ggpairs(
    upper = list(continuous = upper_pearson("#2c7fb8")),
    lower = list(continuous = lower_scatter("#2c7fb8")),
    diag  = list(continuous = wrap("densityDiag", fill = "#a6cee3", colour = "#1f78b4", alpha = 0.7)),
    columnLabels = c("HR_G1", "HR_G2", "HR_G3")
  ) +
  labs(title = "CP34H — Condition G (log₂ PSM, protein level)") +
  theme_bw(base_size = 11)
```

---

## 1.2 *T. pseudonana* (*Thalassiosira pseudonana*)

```{r p1_phyto_load}
phyto_raw <- read_csv("phytoplankton_peptides_filtered.csv", show_col_types = FALSE)
phyto_clean <- phyto_raw %>% filter(!str_detect(`Protein(s)`, contam_pattern))
cat("T. pseudonana peptides — raw:", nrow(phyto_raw), "| after contaminant filter:", nrow(phyto_clean), "\n")

phyto_proteins <- phyto_clean %>%
  mutate(Protein = str_extract(`Protein(s)`, "[^,]+") %>% str_trim()) %>%
  select(Protein, starts_with("PSM Count")) %>%
  group_by(Protein) %>%
  summarise(across(everything(), sum), .groups = "drop") %>%
  rename_with(~ str_remove_all(.x, "PSM Count \\(|\\)"), starts_with("PSM"))

phyto_log <- phyto_proteins %>%
  mutate(across(-Protein, ~ log2(.x + 1))) %>%
  filter(rowSums(select(., -Protein)) > 0)

cat("Protein groups retained:", nrow(phyto_log), "\n")
```

### Condition PEC — control triplicates

```{r p1_phyto_PEC, fig.cap="T. pseudonana — Condition PEC pairwise correlations"}
phyto_log %>%
  select(all_of(c("PEC1", "PEC2", "PEC3"))) %>%
  ggpairs(
    upper = list(continuous = upper_pearson("#2ca25f")),
    lower = list(continuous = lower_scatter("#2ca25f")),
    diag  = list(continuous = wrap("densityDiag", fill = "#b2e2b0", colour = "#2ca25f", alpha = 0.7)),
    columnLabels = c("PEC1", "PEC2", "PEC3")
  ) +
  labs(title = "T. pseudonana — Condition PEC (log₂ PSM, protein level)") +
  theme_bw(base_size = 11)
```

### Condition PED 1:1.5 — triplicates

```{r p1_phyto_PED15, fig.cap="T. pseudonana — Condition PED 1:1.5 pairwise correlations"}
phyto_log %>%
  select(all_of(c("PED1 1.5", "PED2 1.5", "PED3 1.5"))) %>%
  ggpairs(
    upper = list(continuous = upper_pearson("#2ca25f")),
    lower = list(continuous = lower_scatter("#2ca25f")),
    diag  = list(continuous = wrap("densityDiag", fill = "#b2e2b0", colour = "#2ca25f", alpha = 0.7)),
    columnLabels = c("PED1 1.5", "PED2 1.5", "PED3 1.5")
  ) +
  labs(title = "T. pseudonana — Condition PED 1:1.5 (log₂ PSM, protein level)") +
  theme_bw(base_size = 11)
```

### Condition PED 1:2 — replicates 4–5

```{r p1_phyto_PED12, fig.cap="T. pseudonana — Condition PED 1:2 scatter"}
ped12_cols <- c("PED4 1:2", "PED5 1:2")
ped12_data <- phyto_log %>%
  select(all_of(ped12_cols)) %>%
  filter(is.finite(.[[1]]) & is.finite(.[[2]]) & (.[[1]] > 0 | .[[2]] > 0))

r_val <- cor(ped12_data[[1]], ped12_data[[2]], method = "pearson")

ggplot(ped12_data, aes(x = .data[[ped12_cols[1]]], y = .data[[ped12_cols[2]]])) +
  geom_point(alpha = 0.35, size = 1, colour = "#2ca25f") +
  geom_abline(intercept = 0, slope = 1, colour = "firebrick", linewidth = 0.7, linetype = "dashed") +
  annotate("text",
           x = quantile(ped12_data[[1]], 0.15),
           y = quantile(ped12_data[[2]], 0.92),
           label = sprintf("r = %.4f", r_val),
           size = 5, fontface = "bold", colour = "#2ca25f") +
  labs(
    title = "T. pseudonana — Condition PED 1:2 (log₂ PSM, protein level)",
    x = "PED4 1:2  (log₂ PSM)",
    y = "PED5 1:2  (log₂ PSM)"
  ) +
  theme_bw(base_size = 12)
```

---

# Part 2 — NSAF Replicate Scatter (Figs 5C, 6C)

Pairwise log₁₀ NSAF scatter from per-protein NSAF tables. Only proteins with
NSAF > 0 in both replicates of a given pair are plotted. Pearson R² is annotated
on each panel; dashed line = 1:1 identity.

## Shared helpers

```{r p2_helpers, fig.width=13, fig.height=5}
# Contaminant flagging using Protein Description column
contam_patterns <- c(
  "(?i)keratin", "(?i)trypsin", "(?i)albumin|serum albumin",
  "(?i)\\bhomo sapiens\\b", "(?i)\\bbos taurus\\b",
  "(?i)\\bmus musculus\\b|\\brat\\b",
  "(?i)^cRAP|CON__", "(?i)\\(CONTAMINANT\\)"
)

flag_contaminants <- function(df, desc_col = "Protein Description(s)") {
  descs <- df[[desc_col]]
  is_contam <- rep(FALSE, nrow(df))
  for (pat in contam_patterns) is_contam <- is_contam | grepl(pat, descs, perl = TRUE)
  df$contaminant <- is_contam
  df
}

scatter_pair <- function(df, col_x, col_y, label_x, label_y, colour = "#2166AC", title = "") {
  sub <- df %>%
    select(x = all_of(col_x), y = all_of(col_y)) %>%
    filter(x > 0, y > 0) %>%
    mutate(lx = log10(x), ly = log10(y))

  r  <- cor(sub$lx, sub$ly, method = "pearson")
  r2 <- r^2
  n  <- nrow(sub)
  lo <- min(sub$lx, sub$ly) - 0.1
  hi <- max(sub$lx, sub$ly) + 0.1
  ann <- sprintf("R² = %.4f\nn = %s", r2, format(n, big.mark = ","))

  ggplot(sub, aes(lx, ly)) +
    geom_point(size = 0.8, alpha = 0.35, color = colour) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "#888888", linewidth = 0.7) +
    annotate("text", x = lo + 0.05, y = hi - 0.05, label = ann,
             hjust = 0, vjust = 1, size = 3.0, lineheight = 1.4, colour = "#222222") +
    coord_fixed(xlim = c(lo, hi), ylim = c(lo, hi)) +
    labs(title = title,
         x = paste0("log₁₀ NSAF  (", label_x, ")"),
         y = paste0("log₁₀ NSAF  (", label_y, ")")) +
    theme_bw(base_size = 9) +
    theme(plot.title = element_text(face = "bold", size = 9), aspect.ratio = 1)
}
```

## 2.1 CP34H — NSAF scatter

```{r p2_bact_load}
bact_nsaf_raw <- read_csv("Bacteria/proteins_clean.csv", show_col_types = FALSE)
bact_nsaf_raw <- flag_contaminants(bact_nsaf_raw)
bact_nsaf     <- filter(bact_nsaf_raw, !contaminant)
cat("Bacteria:", nrow(bact_nsaf_raw), "raw ->", nrow(bact_nsaf), "after contaminant removal\n")
```

### Control triplicates

```{r p2_bact_ctrl, fig.width=13, fig.height=5}
p1 <- scatter_pair(bact_nsaf, "NSAF (HR_BC1)", "NSAF (HR_BC2)", "BC1", "BC2", "#2166AC", "BC1 vs BC2")
p2 <- scatter_pair(bact_nsaf, "NSAF (HR_BC1)", "NSAF (HR_BC3)", "BC1", "BC3", "#2166AC", "BC1 vs BC3")
p3 <- scatter_pair(bact_nsaf, "NSAF (HR_BC2)", "NSAF (HR_BC3)", "BC2", "BC3", "#2166AC", "BC2 vs BC3")

(p1 | p2 | p3) +
  plot_annotation(title = "Colwellia 34H — Control Triplicates (log₁₀ NSAF)",
                  theme = theme(plot.title = element_text(face = "bold", size = 12)))

ggsave("figures/NSAF_replicate_scatter/BC_Controls_Combined.png", width = 13, height = 5, dpi = 300)
```

### DNA Shield 1:1.5

```{r p2_bact_e, fig.width=13, fig.height=5}
p4 <- scatter_pair(bact_nsaf, "NSAF (HR-E1)", "NSAF (HR_E2)", "E1", "E2", "#4393C3", "E1 vs E2")
p5 <- scatter_pair(bact_nsaf, "NSAF (HR-E1)", "NSAF (HR_E3)", "E1", "E3", "#4393C3", "E1 vs E3")
p6 <- scatter_pair(bact_nsaf, "NSAF (HR_E2)", "NSAF (HR_E3)", "E2", "E3", "#4393C3", "E2 vs E3")

(p4 | p5 | p6) +
  plot_annotation(title = "Colwellia 34H — DNA Shield 1:1.5",
                  theme = theme(plot.title = element_text(face = "bold", size = 12)))

ggsave("figures/NSAF_replicate_scatter/E_Dilutions_Combined.png", width = 13, height = 5, dpi = 300)
```

### DNA Shield 1:1.75

```{r p2_bact_f, fig.width=13, fig.height=5}
p7 <- scatter_pair(bact_nsaf, "NSAF (HR_F1)", "NSAF (HR-F2)", "F1", "F2", "#74ADD1", "F1 vs F2")
p8 <- scatter_pair(bact_nsaf, "NSAF (HR_F1)", "NSAF (HR_F3)", "F1", "F3", "#74ADD1", "F1 vs F3")
p9 <- scatter_pair(bact_nsaf, "NSAF (HR-F2)", "NSAF (HR_F3)", "F2", "F3", "#74ADD1", "F2 vs F3")

(p7 | p8 | p9) +
  plot_annotation(title = "Colwellia 34H — DNA Shield 1:1.75",
                  theme = theme(plot.title = element_text(face = "bold", size = 12)))

ggsave("figures/NSAF_replicate_scatter/F_Dilutions_Combined.png", width = 13, height = 5, dpi = 300)
```

### DNA Shield 1:2

```{r p2_bact_g, fig.width=13, fig.height=5}
p10 <- scatter_pair(bact_nsaf, "NSAF (HR_G1)", "NSAF (HR_G2)", "G1", "G2", "#92C5DE", "G1 vs G2")
p11 <- scatter_pair(bact_nsaf, "NSAF (HR_G1)", "NSAF (HR_G3)", "G1", "G3", "#92C5DE", "G1 vs G3")
p12 <- scatter_pair(bact_nsaf, "NSAF (HR_G2)", "NSAF (HR_G3)", "G2", "G3", "#92C5DE", "G2 vs G3")

(p10 | p11 | p12) +
  plot_annotation(title = "Colwellia 34H — DNA Shield 1:2",
                  theme = theme(plot.title = element_text(face = "bold", size = 12)))

ggsave("figures/NSAF_replicate_scatter/G_Dilutions_Combined.png", width = 13, height = 5, dpi = 300)
```

### All conditions combined — CP34H

```{r p2_bact_all, fig.width=13, fig.height=18}
(p1 | p2 | p3) / (p4 | p5 | p6) / (p7 | p8 | p9) / (p10 | p11 | p12) +
  plot_annotation(
    title    = "Colwellia psychrerythraea 34H — NSAF Replicate Reproducibility",
    subtitle = "All conditions (log₁₀ NSAF, contaminants removed)",
    theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                     plot.subtitle = element_text(size = 10, colour = "grey40"))
  )

ggsave("figures/NSAF_replicate_scatter/Bact_All_Conditions_Combined.png",
       width = 13, height = 18, dpi = 300)
```

### R² summary — CP34H

```{r p2_bact_r2}
pairs_bact <- list(
  list("NSAF (HR_BC1)", "NSAF (HR_BC2)", "BC1 vs BC2", "Control"),
  list("NSAF (HR_BC1)", "NSAF (HR_BC3)", "BC1 vs BC3", "Control"),
  list("NSAF (HR_BC2)", "NSAF (HR_BC3)", "BC2 vs BC3", "Control"),
  list("NSAF (HR-E1)",  "NSAF (HR_E2)",  "E1 vs E2",   "1:1.5"),
  list("NSAF (HR-E1)",  "NSAF (HR_E3)",  "E1 vs E3",   "1:1.5"),
  list("NSAF (HR_E2)",  "NSAF (HR_E3)",  "E2 vs E3",   "1:1.5"),
  list("NSAF (HR_F1)",  "NSAF (HR-F2)",  "F1 vs F2",   "1:1.75"),
  list("NSAF (HR_F1)",  "NSAF (HR_F3)",  "F1 vs F3",   "1:1.75"),
  list("NSAF (HR-F2)",  "NSAF (HR_F3)",  "F2 vs F3",   "1:1.75"),
  list("NSAF (HR_G1)",  "NSAF (HR_G2)",  "G1 vs G2",   "1:2"),
  list("NSAF (HR_G1)",  "NSAF (HR_G3)",  "G1 vs G3",   "1:2"),
  list("NSAF (HR_G2)",  "NSAF (HR_G3)",  "G2 vs G3",   "1:2")
)

r2_bact <- map_dfr(pairs_bact, function(p) {
  sub <- bact_nsaf %>% select(x = all_of(p[[1]]), y = all_of(p[[2]])) %>%
    filter(x > 0, y > 0) %>% mutate(lx = log10(x), ly = log10(y))
  r <- cor(sub$lx, sub$ly)
  tibble(Pair = p[[3]], Condition = p[[4]], r = round(r, 4), R2 = round(r^2, 4), n = nrow(sub))
})

knitr::kable(r2_bact, caption = "CP34H — Pairwise Pearson r and R² (log₁₀ NSAF)")
```

---

## 2.2 *T. pseudonana* — NSAF scatter

```{r p2_phyto_load}
phyto_nsaf_raw <- read_csv("Phytoplankton/proteins_clean.csv", show_col_types = FALSE)
phyto_nsaf_raw <- flag_contaminants(phyto_nsaf_raw)
phyto_nsaf     <- filter(phyto_nsaf_raw, !contaminant)
cat("Phyto:", nrow(phyto_nsaf_raw), "raw ->", nrow(phyto_nsaf), "after contaminant removal\n")
```

### Control triplicates

```{r p2_phyto_ctrl, fig.width=13, fig.height=5}
q1 <- scatter_pair(phyto_nsaf, "NSAF (PEC1)", "NSAF (PEC2)", "PEC1", "PEC2", "#1A7E3B", "PEC1 vs PEC2")
q2 <- scatter_pair(phyto_nsaf, "NSAF (PEC1)", "NSAF (PEC3)", "PEC1", "PEC3", "#1A7E3B", "PEC1 vs PEC3")
q3 <- scatter_pair(phyto_nsaf, "NSAF (PEC2)", "NSAF (PEC3)", "PEC2", "PEC3", "#1A7E3B", "PEC2 vs PEC3")

(q1 | q2 | q3) +
  plot_annotation(title = "T. pseudonana — Control Triplicates (log₁₀ NSAF)",
                  theme = theme(plot.title = element_text(face = "bold", size = 12)))

ggsave("figures/NSAF_replicate_scatter_phyto/PEC_Controls_Combined.png",
       width = 13, height = 5, dpi = 300)
```

### DNA Shield 1:1.5

```{r p2_phyto_15, fig.width=13, fig.height=5}
q4 <- scatter_pair(phyto_nsaf, "NSAF (PED1 1.5)", "NSAF (PED2 1.5)", "PED1", "PED2", "#52B788", "PED1 vs PED2")
q5 <- scatter_pair(phyto_nsaf, "NSAF (PED1 1.5)", "NSAF (PED3 1.5)", "PED1", "PED3", "#52B788", "PED1 vs PED3")
q6 <- scatter_pair(phyto_nsaf, "NSAF (PED2 1.5)", "NSAF (PED3 1.5)", "PED2", "PED3", "#52B788", "PED2 vs PED3")

(q4 | q5 | q6) +
  plot_annotation(title = "T. pseudonana — DNA Shield 1:1.5",
                  theme = theme(plot.title = element_text(face = "bold", size = 12)))

ggsave("figures/NSAF_replicate_scatter_phyto/PED1_5_Dilutions_Combined.png",
       width = 13, height = 5, dpi = 300)
```

### DNA Shield 1:2

```{r p2_phyto_2, fig.width=13, fig.height=5}
q7 <- scatter_pair(phyto_nsaf, "NSAF (PED4 1:2)", "NSAF (PED5 1:2)", "PED4", "PED5", "#95D5B2", "PED4 vs PED5")
q8 <- scatter_pair(phyto_nsaf, "NSAF (PED4 1:2)", "NSAF (PED6 1:2)", "PED4", "PED6", "#95D5B2", "PED4 vs PED6")
q9 <- scatter_pair(phyto_nsaf, "NSAF (PED5 1:2)", "NSAF (PED6 1:2)", "PED5", "PED6", "#95D5B2", "PED5 vs PED6")

(q7 | q8 | q9) +
  plot_annotation(title = "T. pseudonana — DNA Shield 1:2",
                  theme = theme(plot.title = element_text(face = "bold", size = 12)))

ggsave("figures/NSAF_replicate_scatter_phyto/PED4_6_Dilutions_Combined.png",
       width = 13, height = 5, dpi = 300)
```

### All conditions combined — T. pseudonana

```{r p2_phyto_all, fig.width=13, fig.height=14}
(q1 | q2 | q3) / (q4 | q5 | q6) / (q7 | q8 | q9) +
  plot_annotation(
    title    = "Thalassiosira pseudonana — NSAF Replicate Reproducibility",
    subtitle = "All conditions (log₁₀ NSAF, contaminants removed)",
    theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                     plot.subtitle = element_text(size = 10, colour = "grey40"))
  )

ggsave("figures/NSAF_replicate_scatter_phyto/Phyto_All_Conditions_Combined.png",
       width = 13, height = 14, dpi = 300)
```

### R² summary — T. pseudonana

```{r p2_phyto_r2}
pairs_phyto <- list(
  list("NSAF (PEC1)",     "NSAF (PEC2)",     "PEC1 vs PEC2", "Control"),
  list("NSAF (PEC1)",     "NSAF (PEC3)",     "PEC1 vs PEC3", "Control"),
  list("NSAF (PEC2)",     "NSAF (PEC3)",     "PEC2 vs PEC3", "Control"),
  list("NSAF (PED1 1.5)", "NSAF (PED2 1.5)", "PED1 vs PED2", "1:1.5"),
  list("NSAF (PED1 1.5)", "NSAF (PED3 1.5)", "PED1 vs PED3", "1:1.5"),
  list("NSAF (PED2 1.5)", "NSAF (PED3 1.5)", "PED2 vs PED3", "1:1.5"),
  list("NSAF (PED4 1:2)", "NSAF (PED5 1:2)", "PED4 vs PED5", "1:2"),
  list("NSAF (PED4 1:2)", "NSAF (PED6 1:2)", "PED4 vs PED6", "1:2"),
  list("NSAF (PED5 1:2)", "NSAF (PED6 1:2)", "PED5 vs PED6", "1:2")
)

r2_phyto <- map_dfr(pairs_phyto, function(p) {
  sub <- phyto_nsaf %>% select(x = all_of(p[[1]]), y = all_of(p[[2]])) %>%
    filter(x > 0, y > 0) %>% mutate(lx = log10(x), ly = log10(y))
  r <- cor(sub$lx, sub$ly)
  tibble(Pair = p[[3]], Condition = p[[4]], r = round(r, 4), R2 = round(r^2, 4), n = nrow(sub))
})

knitr::kable(r2_phyto, caption = "T. pseudonana — Pairwise Pearson r and R² (log₁₀ NSAF)")
```

---

# Session Info

```{r session}
sessionInfo()
```
