---
title: "DNA Shield Proteomics — Protein-Level Venn Diagrams (Missingness)"
subtitle: "Colwellia psychrerythraea 34H & Thalassiosira pseudonana"
author: "Evelyn Bartlett — Brook Lab"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 6, dpi = 300)
```

## Overview

This document generates protein-level Venn diagrams comparing which proteins
are detected across control and DNA Shield dilution groups for:

- **Bacteria**: CP34H (*Colwellia psychrerythraea* 34H)
  - BC = Control (no DNA Shield)
  - E  = DNA Shield 1:1.5
  - F  = DNA Shield 1:1.75
  - G  = DNA Shield 1:2

- **Phytoplankton**: T. pseudonana (*Thalassiosira pseudonana*)
  - PEC = Control (no DNA Shield)
  - PED 1:1.5 = DNA Shield 1:1.5
  - PED 1:2   = DNA Shield 1:2

Contaminants are removed before any analysis using keyword matching on protein
descriptions (keratins, trypsin, albumin, human/bovine sequences, cRAP tags).
A protein is considered **detected** in a group if it has ≥ 1 PSM in at least
2 of 3 replicates.

---

## Load Packages

```{r packages}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
pkgs <- c("tidyverse", "ggVennDiagram", "ggplot2", "scales", "patchwork")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
}
library(tidyverse)
library(ggVennDiagram)
library(ggplot2)
library(scales)
library(patchwork)
```

---

## Contaminant-Filtering Function

```{r contam_fn}
contaminant_patterns <- c(
  "(?i)keratin",
  "(?i)trypsin",
  "(?i)albumin|serum albumin",
  "(?i)\\bhomo sapiens\\b",
  "(?i)\\bbos taurus\\b",
  "(?i)\\bmus musculus\\b|\\brat\\b",
  "(?i)^cRAP|CON__",
  "(?i)\\(CONTAMINANT\\)"
)

flag_contaminants <- function(df, desc_col = "Protein Description(s)") {
  descs <- df[[desc_col]]
  is_contam <- rep(FALSE, nrow(df))
  for (pat in contaminant_patterns) {
    is_contam <- is_contam | grepl(pat, descs, perl = TRUE)
  }
  df$contaminant <- is_contam
  df
}
```

---

## Helper: Detected Proteins Per Group

A protein is **detected** in a group if PSM > 0 in **≥ 2 of 3 replicates**.

```{r detect_fn}
detected_in_group <- function(df, cols) {
  # cols = character vector of 3 PSM column names
  mat <- df[, cols, drop = FALSE]
  rowSums(mat > 0) >= 2
}
```

---

## Bacteria: CP34H (*Colwellia psychrerythraea* 34H)

### Load & Filter

```{r bact_load}
bact_raw <- read_csv(
  "Bacteria/proteins_clean.csv",
  show_col_types = FALSE
)

cat("Raw protein groups:", nrow(bact_raw), "\n")

bact_raw <- flag_contaminants(bact_raw)
bact_clean <- filter(bact_raw, !contaminant)

cat("After contaminant removal:", nrow(bact_clean), "\n")
cat("Contaminants removed:", sum(bact_raw$contaminant), "\n")
```

### Define Groups

```{r bact_groups}
bact_groups <- list(
  BC  = c("PSMs (HR_BC1)", "PSMs (HR_BC2)", "PSMs (HR_BC3)"),
  E   = c("PSMs (HR-E1)",  "PSMs (HR_E2)",  "PSMs (HR_E3)"),
  F   = c("PSMs (HR_F1)",  "PSMs (HR-F2)",  "PSMs (HR_F3)"),
  G   = c("PSMs (HR_G1)",  "PSMs (HR_G2)",  "PSMs (HR_G3)")
)

group_labels_bact <- c(
  BC = "Control",
  E  = "DNA Shield 1:1.5",
  F  = "DNA Shield 1:1.75",
  G  = "DNA Shield 1:2"
)

# Protein IDs (use first accession in each group string)
prot_ids <- bact_clean$`Protein(s)`

bact_sets <- lapply(names(bact_groups), function(g) {
  detected <- detected_in_group(bact_clean, bact_groups[[g]])
  prot_ids[detected]
})
names(bact_sets) <- group_labels_bact

# Summary
bact_counts <- sapply(bact_sets, length)
print(bact_counts)
```

### Venn Diagram — All Four Groups

```{r bact_venn_4, fig.width=9, fig.height=8}
# Bacteria colour palette (blue family, per poster colour scheme)
bact_colours <- c("#2166AC", "#4393C3", "#92C5DE", "#D1E5F0")

p_bact_4 <- ggVennDiagram(
  bact_sets,
  label       = "count",
  label_alpha = 0,
  set_color   = bact_colours
) +
  scale_fill_gradient(low = "#EFF5FB", high = "#2166AC") +
  scale_color_manual(values = bact_colours) +
  labs(
    title    = expression(italic("Colwellia psychrerythraea")~"34H — Protein Detection"),
    subtitle = "Proteins detected (≥ 1 PSM in ≥ 2/3 replicates); contaminants removed",
    fill     = "Proteins (n)"
  ) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    legend.position = "right"
  )

print(p_bact_4)

dir.create("figures/venn_diagrams", recursive = TRUE, showWarnings = FALSE)
ggsave("figures/venn_diagrams/Bact_protein_venn_4group.png",
       p_bact_4, width = 9, height = 8, dpi = 300)
ggsave("figures/venn_diagrams/Bact_protein_venn_4group.pdf",
       p_bact_4, width = 9, height = 8)
cat("Saved: figures/venn_diagrams/Bact_protein_venn_4group.png\n")
```

### Venn Diagram — Control vs. Best Dilution (1:1.5)

```{r bact_venn_2, fig.width=7, fig.height=6}
bact_sets_2 <- bact_sets[c("Control", "DNA Shield 1:1.5")]

p_bact_2 <- ggVennDiagram(
  bact_sets_2,
  label       = "count",
  label_alpha = 0,
  set_color   = bact_colours[1:2]
) +
  scale_fill_gradient(low = "#EFF5FB", high = "#2166AC") +
  scale_color_manual(values = bact_colours[1:2]) +
  labs(
    title    = expression(italic("Colwellia")~"34H — Control vs. DNA Shield 1:1.5"),
    subtitle = "Proteins detected (≥ 1 PSM in ≥ 2/3 replicates); contaminants removed",
    fill     = "Proteins (n)"
  ) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "grey40")
  )

print(p_bact_2)

ggsave("figures/venn_diagrams/Bact_protein_venn_ctrl_vs_1.5.png",
       p_bact_2, width = 7, height = 6, dpi = 300)
ggsave("figures/venn_diagrams/Bact_protein_venn_ctrl_vs_1.5.pdf",
       p_bact_2, width = 7, height = 6)
cat("Saved: figures/venn_diagrams/Bact_protein_venn_ctrl_vs_1.5.png\n")
```

### Dilution Trend — Proteins Unique to Control

```{r bact_unique_trend}
# How many control-only proteins are recovered as dilution increases?
ctrl_only <- setdiff(bact_sets[["Control"]],
                     Reduce(union, bact_sets[c("DNA Shield 1:1.5",
                                               "DNA Shield 1:1.75",
                                               "DNA Shield 1:2")]))

trend_df <- tibble(
  Group     = names(bact_sets)[-1],
  Dilution  = c("1:1.5", "1:1.75", "1:2"),
  N_detected = sapply(bact_sets[-1], length),
  N_shared_with_ctrl = sapply(bact_sets[-1], function(s) length(intersect(s, bact_sets[["Control"]]))),
  N_ctrl_only = sapply(bact_sets[-1], function(s) length(setdiff(bact_sets[["Control"]], s)))
)

print(trend_df)
write_csv(trend_df, "figures/venn_diagrams/Bact_dilution_trend.csv")
```

---

## Phytoplankton: T. pseudonana (*Thalassiosira pseudonana*)

### Load & Filter

```{r phyto_load}
phyto_raw <- read_csv(
  "Phytoplankton/proteins_clean.csv",
  show_col_types = FALSE
)

cat("Raw protein groups:", nrow(phyto_raw), "\n")

phyto_raw <- flag_contaminants(phyto_raw)
phyto_clean <- filter(phyto_raw, !contaminant)

cat("After contaminant removal:", nrow(phyto_clean), "\n")
cat("Contaminants removed:", sum(phyto_raw$contaminant), "\n")
```

### Define Groups

```{r phyto_groups}
phyto_groups <- list(
  PEC     = c("PSMs (PEC1)",       "PSMs (PEC2)",       "PSMs (PEC3)"),
  PED_1.5 = c("PSMs (PED1 1.5)",   "PSMs (PED2 1.5)",   "PSMs (PED3 1.5)"),
  PED_2   = c("PSMs (PED4 1:2)",   "PSMs (PED5 1:2)",   "PSMs (PED6 1:2)")
)

group_labels_phyto <- c(
  PEC     = "Control",
  PED_1.5 = "DNA Shield 1:1.5",
  PED_2   = "DNA Shield 1:2"
)

phyto_ids <- phyto_clean$`Protein(s)`

phyto_sets <- lapply(names(phyto_groups), function(g) {
  detected <- detected_in_group(phyto_clean, phyto_groups[[g]])
  phyto_ids[detected]
})
names(phyto_sets) <- group_labels_phyto

phyto_counts <- sapply(phyto_sets, length)
print(phyto_counts)
```

### Venn Diagram — All Three Groups

```{r phyto_venn_3, fig.width=8, fig.height=7}
# Phytoplankton colour palette (green family, per poster colour scheme)
phyto_colours <- c("#1A7E3B", "#52B788", "#B7E4C7")

p_phyto_3 <- ggVennDiagram(
  phyto_sets,
  label       = "count",
  label_alpha = 0,
  set_color   = phyto_colours
) +
  scale_fill_gradient(low = "#F0FAF3", high = "#1A7E3B") +
  scale_color_manual(values = phyto_colours) +
  labs(
    title    = expression(italic("Thalassiosira pseudonana")~"— Protein Detection"),
    subtitle = "Proteins detected (≥ 1 PSM in ≥ 2/3 replicates); contaminants removed",
    fill     = "Proteins (n)"
  ) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    legend.position = "right"
  )

print(p_phyto_3)

ggsave("figures/venn_diagrams/Phyto_protein_venn_3group.png",
       p_phyto_3, width = 8, height = 7, dpi = 300)
ggsave("figures/venn_diagrams/Phyto_protein_venn_3group.pdf",
       p_phyto_3, width = 8, height = 7)
cat("Saved: figures/venn_diagrams/Phyto_protein_venn_3group.png\n")
```

### Venn Diagram — Control vs. Each Dilution

```{r phyto_venn_2, fig.width=12, fig.height=5}
p_phyto_2a <- ggVennDiagram(
  phyto_sets[c("Control", "DNA Shield 1:1.5")],
  label = "count", label_alpha = 0,
  set_color = phyto_colours[1:2]
) +
  scale_fill_gradient(low = "#F0FAF3", high = "#1A7E3B") +
  scale_color_manual(values = phyto_colours[1:2]) +
  labs(title = "Control vs. 1:1.5", fill = "n") +
  theme(plot.title = element_text(face = "bold", size = 11))

p_phyto_2b <- ggVennDiagram(
  phyto_sets[c("Control", "DNA Shield 1:2")],
  label = "count", label_alpha = 0,
  set_color = phyto_colours[c(1,3)]
) +
  scale_fill_gradient(low = "#F0FAF3", high = "#1A7E3B") +
  scale_color_manual(values = phyto_colours[c(1,3)]) +
  labs(title = "Control vs. 1:2", fill = "n") +
  theme(plot.title = element_text(face = "bold", size = 11))

p_phyto_combined <- p_phyto_2a + p_phyto_2b +
  plot_annotation(
    title    = expression(italic("Thalassiosira pseudonana")~"— Control vs. Each Dilution"),
    subtitle = "Proteins detected (≥ 1 PSM in ≥ 2/3 replicates); contaminants removed",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10, color = "grey40")
    )
  )

print(p_phyto_combined)

ggsave("figures/venn_diagrams/Phyto_protein_venn_ctrl_vs_dilutions.png",
       p_phyto_combined, width = 12, height = 5, dpi = 300)
ggsave("figures/venn_diagrams/Phyto_protein_venn_ctrl_vs_dilutions.pdf",
       p_phyto_combined, width = 12, height = 5)
cat("Saved: figures/venn_diagrams/Phyto_protein_venn_ctrl_vs_dilutions.png\n")
```

---

## Summary Table

```{r summary_table}
summary_df <- bind_rows(
  tibble(
    Organism  = "Colwellia 34H",
    Group     = names(bact_sets),
    N_proteins = sapply(bact_sets, length)
  ),
  tibble(
    Organism  = "T. pseudonana",
    Group     = names(phyto_sets),
    N_proteins = sapply(phyto_sets, length)
  )
)

knitr::kable(summary_df, caption = "Protein counts per group after contaminant removal")
```

---

## Session Info

```{r session}
sessionInfo()
```
