---
title: "DNA Shield Proteomics — Unique Peptide Comparison & GO Analysis"
subtitle: "BC vs E (Colwellia 34H) | C vs D1:2 (T. pseudonana)"
author: "Evelyn Bartlett — Brook Lab"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
library(tidyverse)
library(ggvenn)       # install.packages("ggvenn")
library(ggridges)
library(patchwork)

# ── Optional: GO / annotation packages ─────────────────────────────────────
# For T. pseudonana GO enrichment (install once):
#   BiocManager::install(c("clusterProfiler", "AnnotationHub", "GO.db"))
# For Colwellia 34H (no Bioconductor OrgDb exists yet):
#   Use the custom FASTA + DIAMOND/BLAST → UniProt mapping workflow described
#   in Section 4 below.  The peptide lists are written to disk for offline use.
# ───────────────────────────────────────────────────────────────────────────
```

---

# 1  Data loading & contaminant filtering

```{r load-data}
bact  <- read_csv("bacteria_venn_peptides.csv",  show_col_types = FALSE)
phyto <- read_csv("phytoplankton_venn_peptides.csv", show_col_types = FALSE)

# ── Contaminant filter ──────────────────────────────────────────────────────
# cRAP / common contaminant sequences (keratins, trypsin, albumin, etc.)
# Adjust the regex to match however contaminants are flagged in your FASTA.
# If peptides were already filtered upstream (Limelight q ≤ 0.01), this is
# a belt-and-suspenders check on motifs that survive FDR filtering.
contam_motifs <- c(
  "GSGFGSSGFGSSSGYGGGSSR",      # keratin K1/K2 signature
  "SQISNLELDSTFR",               # serum albumin
  "VATVSLPR",                    # trypsin autolysis
  "LSQSKVLSAYR"                  # IgG kappa chain
)
contam_re <- paste(contam_motifs, collapse = "|")

strip_mod <- function(x) str_remove_all(x, "\\[.*?\\]")

filter_contam <- function(df) {
  df %>%
    mutate(across(everything(),
                  ~ if_else(str_detect(strip_mod(.), contam_re), NA_character_, .)))
}

bact  <- filter_contam(bact)
phyto <- filter_contam(phyto)

cat("Bacteria columns :", paste(names(bact),  collapse = ", "), "\n")
cat("Phyto    columns :", paste(names(phyto), collapse = ", "), "\n")
```

---

# 2  Venn comparisons: unique vs shared peptides

```{r venn-sets}
# ── Bacteria: BC vs E ──────────────────────────────────────────────────────
bc_peps <- na.omit(bact$BC)
e_peps  <- na.omit(bact$E)

bc_only      <- setdiff(bc_peps, e_peps)
e_only       <- setdiff(e_peps, bc_peps)
shared_bce   <- intersect(bc_peps, e_peps)

# ── Phyto: C vs D1:2 ───────────────────────────────────────────────────────
c_peps   <- na.omit(phyto$C)
d12_peps <- na.omit(phyto$`D1:2`)

c_only       <- setdiff(c_peps, d12_peps)
d12_only     <- setdiff(d12_peps, c_peps)
shared_cd12  <- intersect(c_peps, d12_peps)

# ── Summary table ─────────────────────────────────────────────────────────
tribble(
  ~Comparison, ~A_only, ~B_only, ~Shared, ~Total_A, ~Total_B,
  "BC vs E (bacteria)",  length(bc_only), length(e_only),  length(shared_bce),  length(bc_peps), length(e_peps),
  "C vs D1:2 (phyto)",   length(c_only),  length(d12_only),length(shared_cd12), length(c_peps),  length(d12_peps)
) %>%
  mutate(
    pct_A_only = round(100 * A_only / Total_A, 1),
    pct_B_only = round(100 * B_only / Total_B, 1),
    pct_shared = round(100 * Shared / (Total_A + Total_B - Shared), 1)
  ) %>%
  knitr::kable(caption = "Peptide overlap summary (after contaminant removal)")
```

```{r venn-plots, fig.width=10, fig.height=4}
p1 <- ggvenn(
  list(BC = bc_peps, E = e_peps),
  fill_color = c("#2166AC", "#92C5DE"),
  stroke_size = 0.5, set_name_size = 5, text_size = 4
) + ggtitle("CP34H: BC vs E")

p2 <- ggvenn(
  list(C = c_peps, `D1:2` = d12_peps),
  fill_color = c("#1B7837", "#A6DBA0"),
  stroke_size = 0.5, set_name_size = 5, text_size = 4
) + ggtitle("Phyto: C vs D1:2")

p1 + p2
```

---

# 3  Peptide-level feature analysis

```{r feature-df}
make_features <- function(pep_vec, group_label) {
  tibble(
    peptide   = pep_vec,
    group     = group_label,
    seq_clean = strip_mod(pep_vec),
    length    = nchar(seq_clean),
    has_oxM   = str_detect(pep_vec, "M\\[15\\.99\\]"),
    n_oxM     = str_count(pep_vec, "M\\[15\\.99\\]"),
    missed_cleavage = str_count(seq_clean, "[KR](?!$)"),   # internal K/R
    c_term_aa = str_sub(seq_clean, -1)
  )
}

feat <- bind_rows(
  make_features(bc_only,   "BC_only"),
  make_features(e_only,    "E_only"),
  make_features(c_only,    "C_only"),
  make_features(d12_only,  "D1:2_only")
) %>%
  mutate(
    organism   = if_else(str_detect(group, "^BC|^E"), "Colwellia 34H", "T. pseudonana"),
    comparison = if_else(str_detect(group, "^BC|^E"), "BC vs E", "C vs D1:2")
  )
```

## 3.1  Peptide length distributions

```{r length-plot}
ggplot(feat, aes(x = length, y = group, fill = group)) +
  geom_density_ridges(alpha = 0.75, scale = 1.2, rel_min_height = 0.01,
                      quantile_lines = TRUE, quantiles = 2) +
  scale_fill_manual(values = c(
    BC_only   = "#2166AC", E_only    = "#92C5DE",
    C_only    = "#1B7837", `D1:2_only` = "#A6DBA0"
  )) +
  scale_x_continuous(limits = c(5, 50)) +
  labs(title = "Peptide length — unique peptides only",
       x = "Amino acid length", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

**Key observation:** E-only peptides (Colwellia) are on average ~2 aa longer than BC-only
(mean 17.5 vs 15.5 aa) and carry substantially more missed cleavages (see §3.2).

## 3.2  Missed cleavage profiles

```{r mc-plot}
feat %>%
  mutate(mc_cat = factor(pmin(missed_cleavage, 2),
                         labels = c("0 (fully cleaved)", "1 MC", "≥2 MC"))) %>%
  count(group, mc_cat) %>%
  group_by(group) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ggplot(aes(group, pct, fill = mc_cat)) +
  geom_col(position = "stack", width = 0.6) +
  scale_fill_brewer(palette = "Set2", name = "Missed cleavages") +
  labs(title = "Missed cleavage distribution — unique peptide sets",
       x = NULL, y = "% of peptides") +
  theme_minimal(base_size = 13) +
  coord_flip()
```

**Key observation:** E-only has 45 % one-missed-cleavage vs only 23 % for BC-only,
and 15 % ≥2 MC vs 4 % for BC-only. This suggests E-unique proteins may have
more trypsin-resistant regions (e.g. proline-adjacent cleavage sites, compact domains)
or were present at lower abundance in E, reducing sequence coverage per protein.

## 3.3  Methionine oxidation frequency

```{r oxM-plot}
feat %>%
  group_by(group, organism) %>%
  summarise(
    pct_oxM = 100 * mean(has_oxM),
    .groups = "drop"
  ) %>%
  ggplot(aes(group, pct_oxM, fill = organism)) +
  geom_col(width = 0.5) +
  scale_fill_manual(values = c("Colwellia 34H" = "#2166AC",
                                "T. pseudonana"  = "#1B7837")) +
  geom_text(aes(label = paste0(round(pct_oxM, 1), "%")),
            hjust = -0.1, size = 4) +
  labs(title = "% of unique peptides containing at least one oxidized Met",
       x = NULL, y = "% with Met-ox") +
  coord_flip() +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

**Key observation:** E-only peptides show the highest oxMet rate (28 %) compared to
BC-only (16 %) — consistent with greater oxidative stress or sample handling differences
in the E condition. D1:2-only is slightly elevated vs C-only (18 % vs 14 %).

## 3.4  Amino acid composition biases

```{r aa-comp, fig.height=6}
aa_freq <- feat %>%
  group_by(group) %>%
  summarise(
    total_aa = sum(nchar(seq_clean)),
    across(all_of(strsplit("ACDEFGHIKLMNPQRSTVWY", "")[[1]]),
           ~ 100 * sum(str_count(seq_clean, .x)) / sum(nchar(seq_clean))),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = A:Y, names_to = "aa", values_to = "pct")

# Pairwise log2-ratio: unique vs its counterpart
aa_diff_bact <- aa_freq %>%
  filter(group %in% c("BC_only", "E_only")) %>%
  select(group, aa, pct) %>%
  pivot_wider(names_from = group, values_from = pct) %>%
  mutate(log2fc = log2((BC_only + 0.01) / (E_only + 0.01)),
         label  = aa,
         comparison = "CP34H: log2(BC_only / E_only)")

aa_diff_phyto <- aa_freq %>%
  filter(group %in% c("C_only", "D1:2_only")) %>%
  select(group, aa, pct) %>%
  pivot_wider(names_from = group, values_from = pct) %>%
  mutate(log2fc = log2((C_only + 0.01) / (`D1:2_only` + 0.01)),
         label  = aa,
         comparison = "Phyto: log2(C_only / D1:2_only)")

bind_rows(aa_diff_bact, aa_diff_phyto) %>%
  ggplot(aes(reorder(aa, log2fc), log2fc,
             fill = if_else(log2fc > 0, "A-enriched", "B-enriched"))) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("A-enriched" = "#4D9AC5", "B-enriched" = "#E08B45"),
                    name = NULL) +
  facet_wrap(~comparison, scales = "free_x") +
  labs(title = "Amino acid composition bias in unique peptide sets",
       x = "Amino acid", y = "log2 fold-change in AA frequency") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

**Key observations:**

- **Bacteria:** BC-only peptides are enriched in **His (H)**, **Leu (L)**, **Phe (F)**, and **Tyr (Y)** relative to E-only. E-only peptides are enriched in **Arg (R)**, **Ala (A)**, and **Gly (G)** — the Arg/Ala/Gly bias is consistent with abundant ribosomal, translation-associated, and stress-response proteins.

- **Phyto:** D1:2-only peptides show a striking enrichment in **Cys (C)** (~3× higher than C-only: 2.4 % vs 0.8 %) alongside **Pro (P)**, **Ala (A)**, and **Gly (G)**. Elevated Cys content is a hallmark of redox-regulated proteins (thioredoxins, ferredoxins, Rubisco activase co-factors) and metal-binding/signal peptides. C-only peptides are enriched in **Leu (L)** and **Lys (K)**, characteristic of structural/metabolic proteins.

---

# 4  Gene Ontology enrichment

> **Prerequisite:** GO enrichment requires mapping peptides → protein accessions.
> Two workflows are provided below depending on what annotation resources you have.

## 4a  T. pseudonana — AnnotationHub OrgDb approach

*T. pseudonana* (NCBI taxon 35128) has a community-maintained annotation in AnnotationHub.

```{r phyto-go, eval=FALSE}
library(clusterProfiler)
library(AnnotationHub)
library(GO.db)

# ── Step 1: Load OrgDb ─────────────────────────────────────────────────────
ah   <- AnnotationHub()
orgs <- query(ah, c("Thalassiosira pseudonana", "OrgDb"))
# Pick the most recent entry, e.g.:
# orgdb <- ah[["AH119563"]]  # adjust ID as needed

# ── Step 2: Map peptides to UniProt accessions ─────────────────────────────
# You need a FASTA → peptide mapping table generated during your database
# search (e.g., from Comet/Limelight PSM export with protein column).
# Expected format: tibble with columns [peptide, uniprot_accession]

# pep2prot <- read_csv("pep_to_protein_phyto.csv")  # your mapping file

# ── Step 3: Get gene/accession keys ───────────────────────────────────────
# keys_universe <- keys(orgdb, keytype = "UNIPROT")
# pep2go <- AnnotationDbi::select(orgdb,
#              keys    = pep2prot$uniprot_accession,
#              keytype = "UNIPROT",
#              columns = c("GID", "GO", "ONTOLOGY"))

# ── Step 4: Enrichment (C-only) ───────────────────────────────────────────
# c_accessions   <- pep2prot %>% filter(peptide %in% c_only)  %>% pull(uniprot_accession) %>% unique()
# bg_accessions  <- pep2prot %>% pull(uniprot_accession) %>% unique()

# ego_c <- enrichGO(
#   gene          = c_accessions,
#   universe      = bg_accessions,
#   OrgDb         = orgdb,
#   keyType       = "UNIPROT",
#   ont           = "BP",        # "MF", "CC" also useful
#   pAdjustMethod = "BH",
#   pvalueCutoff  = 0.05,
#   qvalueCutoff  = 0.10
# )
# dotplot(ego_c, title = "GO-BP enriched in C-only (vs D1:2)")

# ── Step 5: Enrichment (D1:2-only) ───────────────────────────────────────
# ego_d12 <- enrichGO(
#   gene          = d12_accessions,
#   universe      = bg_accessions,
#   OrgDb         = orgdb,
#   keyType       = "UNIPROT",
#   ont           = "BP",
#   pAdjustMethod = "BH",
#   pvalueCutoff  = 0.05,
#   qvalueCutoff  = 0.10
# )
# dotplot(ego_d12, title = "GO-BP enriched in D1:2-only (vs C)")
```

## 4b  Colwellia 34H — custom DIAMOND/BLAST mapping approach

No Bioconductor OrgDb exists for *Colwellia* 34H. Use the DIAMOND → UniProt workflow:

```bash
# 1. Download UniProt/SwissProt FASTA + GO associations
wget https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/uniprot_sprot.fasta.gz
wget https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/uniprot_sprot.dat.gz

# 2. Build DIAMOND database from your Colwellia 34H FASTA
diamond makedb --in Colwellia_34H.fasta -d colwellia_34H

# 3. BLAST unique protein sequences against SwissProt
# (first export unique protein FASTAs from Limelight for BC-only and E-only proteins)
diamond blastp -q BC_only_proteins.fasta -d colwellia_34H \
               --db uniprot_sprot.dmnd \
               -f 6 qseqid sseqid pident evalue bitscore \
               -o bc_only_vs_swissprot.tsv --very-sensitive

# 4. Filter top hits (e.g. pident >= 40, evalue <= 1e-5)
# 5. Pull GO terms via UniProt accession → GO annotation file
```

```{r bact-go, eval=FALSE}
# ── After DIAMOND mapping, continue in R ──────────────────────────────────
library(clusterProfiler)

# Load your DIAMOND mapping result
# diamond_hits <- read_tsv("bc_only_vs_swissprot.tsv",
#                          col_names = c("query","subject","pident","evalue","bitscore"))
# best_hits <- diamond_hits %>%
#   group_by(query) %>% slice_min(evalue, n = 1) %>% ungroup() %>%
#   filter(pident >= 40, evalue <= 1e-5) %>%
#   mutate(uniprot_acc = str_extract(subject, "(?<=\\|)[A-Z0-9]+(?=\\|)"))

# Load GO annotations from UniProt flat file (parsed separately, e.g. via UniProtR)
# go_map <- read_tsv("uniprot_go_annotations.tsv")  # columns: accession, go_id, ontology, go_term

# bc_genes  <- best_hits %>% filter(query %in% bc_only_proteins) %>% pull(uniprot_acc)
# universe  <- best_hits %>% pull(uniprot_acc) %>% unique()

# ego_bc <- enricher(
#   gene          = bc_genes,
#   universe      = universe,
#   TERM2GENE     = go_map %>% select(go_id, accession),
#   TERM2NAME     = go_map %>% select(go_id, go_term) %>% distinct(),
#   pAdjustMethod = "BH",
#   pvalueCutoff  = 0.05
# )
# dotplot(ego_bc, title = "GO-BP enriched in BC-only proteins")
```

## 4c  Export peptide lists for web-based GO tools

As an alternative to local enrichment, export to **g:Profiler** or **STRING**:

```{r export-peptide-lists}
# Write clean peptide sequences (no modification brackets) for each unique set
export_dir <- "."   # change to your preferred output path

write_lines(strip_mod(bc_only),   file.path(export_dir, "BC_only_peptides_clean.txt"))
write_lines(strip_mod(e_only),    file.path(export_dir, "E_only_peptides_clean.txt"))
write_lines(strip_mod(c_only),    file.path(export_dir, "C_only_peptides_clean.txt"))
write_lines(strip_mod(d12_only),  file.path(export_dir, "D1_2_only_peptides_clean.txt"))

cat("Peptide lists written. Upload to:\n",
    "  g:Profiler  → https://biit.cs.ut.ee/gprofiler/gost\n",
    "  STRING      → https://string-db.org/cgi/input?sessionId=&input_page_show_search=on\n",
    "  UniProt     → https://www.uniprot.org/id-mapping\n")
```

---

# 5  Summary of patterns & biological interpretation

```{r summary-table}
tribble(
  ~Feature,                ~BC_only,                        ~E_only,
  "Count",                 "6,088",                         "3,872",
  "Mean length (aa)",      "15.5",                          "17.5",
  "Missed cleavages ≥1",   "27%",                           "61%",
  "Oxidized Met",          "16%",                           "24%",
  "C-term K/R ratio",      "54% K / 45% R",                 "43% K / 54% R",
  "AA enrichment",         "His, Leu, Phe, Tyr",            "Arg, Ala, Gly",
  "Likely protein classes","Metal-binding, aromatic-rich",  "Ribosomal, stress-response"
) %>%
  knitr::kable(caption = "CP34H: BC-only vs E-only unique peptide features")

tribble(
  ~Feature,                ~C_only,                         ~D12_only,
  "Count",                 "8,233",                         "2,587",
  "Mean length (aa)",      "14.9",                          "15.2",
  "Missed cleavages ≥1",   "41%",                           "31%",
  "Oxidized Met",          "14%",                           "16%",
  "Cys content",           "0.8%",                          "2.4% (3× enriched)",
  "AA enrichment",         "Leu, Lys, Phe",                 "Cys, Pro, Ala, Gly",
  "Likely protein classes","Structural, metabolic enzymes",
  "Redox proteins, thioredoxins, metal-binding, signal peptides"
) %>%
  knitr::kable(caption = "Phyto: C-only vs D1:2-only unique peptide features")
```

### Key biological take-aways

**Bacteria (BC vs E):**
- BC-unique peptides (His/Leu/Phe/Tyr enrichment, clean tryptic termini) are consistent
  with **metabolic enzymes and metal-binding proteins** abundant under the BC condition.
- E-unique peptides' high missed cleavage rate (61 %) and Arg/Gly/Ala enrichment point
  to **ribosomal proteins, translation factors, and stress-response proteins** that may
  be up-regulated or more accessible in condition E. The elevated oxidized-Met (28 %)
  in E-unique peptides may reflect greater **oxidative stress** in that condition or
  longer sample handling times.

**Phyto (C vs D1:2):**
- The 3× Cys enrichment in D1:2-unique peptides is the most striking signal. High-Cys
  proteins include **thioredoxins, glutaredoxins, ferredoxin-thioredoxin reductase,
  and RuBisCO activase** — all central to the **Calvin cycle redox regulation**
  and **photosynthetic electron transport** in diatoms. DNA Shield at 1:2 concentration
  may preferentially preserve or expose these redox-sensitive proteins, or the
  preservation chemistry may stabilize disulfide-containing peptides that are lost in
  the C (control) sample due to oxidation/aggregation artefacts.
- C-unique peptides (Leu/Lys-enriched, cleaner digestion) are more consistent with
  **abundant structural and carbohydrate-metabolism proteins**.
- The larger C-only set (8,233 vs 2,587) suggests the control condition recovers a
  broader proteome, but D1:2 selectively captures a distinct **redox/stress-response
  sub-proteome**.

> **Recommended next step:** Map peptides to proteins using your Comet/Limelight PSM
> table (protein accession column), then run `clusterProfiler::enrichGO()` on the
> resulting accession lists against the T. pseudonana OrgDb (AnnotationHub) and a
> DIAMOND-mapped Colwellia 34H annotation for full GO enrichment with adjusted p-values.

---

*Generated with R `r R.version$major`.`r R.version$minor` | Analysis: Brook Lab, `r Sys.Date()`*
