---
title: "GO-Annotated Volcano Plots — DNA Shield Proteomics"
author: "Evelyn Bartlett (Brook Lab)"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    fig_width: 8
    fig_height: 7
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile,
                      encoding = encoding,
                      knit_root_dir = getwd()) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  dpi     = 300
)
```

## Overview

Differential abundance volcano plots comparing:

1. **CP34H** (*Colwellia psychrerythraea* 34H) — B-CTL vs B-1.5 (DNA Shield 1:1.5)
2. **T. pseudonana** (*Thalassiosira pseudonana*) — P-CTL vs P-2.0 (DNA Shield 1:2)

Points are colored by GO biological process category, assigned via priority-ordered
regex matching against metaGOmics2 annotations. Statistics: Welch's t-test on
log₁₀ NSAF values per protein, with Benjamini-Hochberg FDR correction. Significance
threshold: FDR < 0.05 AND |log₂FC| > 1.

Upstream pipeline: Comet 2023.02 → Percolator via Limelight; PSM- and peptide-level
q ≤ 0.01 FDR; contaminants excluded.

---

## Packages

```{r packages}
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(arrow)    # arrow::read_parquet()
library(scales)
```

---

## GO category definitions

Priority order: the first matching category wins.

```{r go-categories}
GO_CATS <- list(
  list(name    = "Photosynthesis",
       color   = "#2ecc71",
       pattern = "photosynthes|light reaction|chlorophyll|photosystem|calvin"),
  list(name    = "Translation / Ribosome",
       color   = "#e74c3c",
       pattern = "translat|ribosom|trna|aminoacyl-trna|peptide biosynthes|rrna"),
  list(name    = "DNA / RNA processes",
       color   = "#3498db",
       pattern = "dna replic|dna repair|transcription|rna process|nucleotide|chromosome|dna metabol"),
  list(name    = "Protein quality control",
       color   = "#9b59b6",
       pattern = "protein fold|chaperone|proteolysis|proteasome|ubiquitin|protein catabol|misfolded"),
  list(name    = "Metabolic process",
       color   = "#f39c12",
       pattern = "biosynthes|catabolic|metabolic process|carbon|nitrogen|lipid|fatty acid|glucos|glycol|amino acid|carbohydrate"),
  list(name    = "Transport",
       color   = "#1abc9c",
       pattern = "transport|transmembrane transport|ion transport|export|import|secretion"),
  list(name    = "Stress response",
       color   = "#e67e22",
       pattern = "response to stress|response to heat|response to cold|response to oxidat|response to osmot|stress"),
  list(name    = "Signal / Regulation",
       color   = "#e91e63",
       pattern = "signal transduct|regulation|chemotaxis|phosphorylat|two-component"),
  list(name    = "Cell cycle / Division",
       color   = "#607d8b",
       pattern = "cell cycle|cell division|cytokinesis|chromosome segregat")
)

CAT_OTHER       <- list(name = "Other / Unassigned", color = "#aaaaaa")
CAT_HYPOTHETICAL <- list(name = "Hypothetical",       color = "#dddddd")
```

---

## Helper functions

### Build protein → GO name map from metaGOmics2 output

```{r fn-load-go-map}
#' Build a protein-to-BP-GO-name map from one or more metaGOmics2 list dirs.
#'
#' @param list_dirs Character vector of paths to list_XXX folders, each
#'   containing go_terms.csv and peptide_mapping.parquet.
#' @return Named list: protein_id → character vector of lowercase BP GO names.
load_go_map <- function(list_dirs) {
  prot_go <- list()   # protein_id → set of BP GO term names

  for (d in list_dirs) {
    # GO term metadata: go_id, name, namespace
    gt <- read_csv(file.path(d, "go_terms.csv"), show_col_types = FALSE)
    bp_info <- gt %>%
      filter(namespace == "biological_process") %>%
      select(go_id, name) %>%
      mutate(name = tolower(name))

    # Peptide → GO mapping
    pm <- arrow::read_parquet(file.path(d, "peptide_mapping.parquet"))

    # peptide_go_terms may be a list column (Arrow list type) or a string
    # (Python-style repr e.g. "['GO:0006810', 'GO:0055085']")
    if (is.list(pm$peptide_go_terms)) {
      # Already a list column — each element is a character vector of GO IDs
      go_lists <- pm$peptide_go_terms
    } else {
      # String representation — parse with eval(parse())
      go_lists <- lapply(pm$peptide_go_terms, function(s) {
        if (is.na(s) || s == "") return(character(0))
        tryCatch(eval(parse(text = s)), error = function(e) character(0))
      })
    }

    for (i in seq_len(nrow(pm))) {
      pid  <- pm$background_protein[i]
      gos  <- go_lists[[i]]
      if (length(gos) == 0) next

      # Keep only BP terms
      bp_names <- bp_info$name[bp_info$go_id %in% gos]
      if (length(bp_names) == 0) next

      prot_go[[pid]] <- unique(c(prot_go[[pid]], bp_names))
    }
  }

  message("GO map built: ", length(prot_go), " proteins annotated")
  prot_go
}
```

### Assign GO category to a protein group

```{r fn-assign-go-category}
#' Assign the highest-priority GO category to a protein group entry.
#'
#' @param protein_ids_str Comma-separated protein IDs (from "Protein(s)" column).
#' @param prot_go_map Named list from load_go_map().
#' @return List with $name and $color.
assign_go_category <- function(protein_ids_str, prot_go_map) {
  ids    <- trimws(strsplit(as.character(protein_ids_str), ",")[[1]])
  all_bp <- unique(unlist(prot_go_map[ids[ids %in% names(prot_go_map)]]))

  if (length(all_bp) == 0) return(CAT_HYPOTHETICAL)

  combined <- paste(all_bp, collapse = " ")
  for (cat in GO_CATS) {
    if (grepl(cat$pattern, combined, ignore.case = TRUE)) return(cat)
  }
  CAT_OTHER
}
```

### Clean protein description label

```{r fn-clean-label}
#' Strip UniProt annotation noise and truncate for plot labels.
clean_label <- function(desc_str, max_chars = 38) {
  raw <- strsplit(as.character(desc_str), ",")[[1]][1]
  lab <- gsub("\\[.*?\\]|RecName: Full=|AltName:.*|;.*", "", raw)
  lab <- trimws(lab)
  substr(lab, 1, max_chars)
}
```

### Core volcano plot function

```{r fn-run-volcano}
#' Run a full GO-annotated volcano analysis and return a list with the
#' annotated data frame and the ggplot object.
#'
#' @param nsaf_path   Path to the tab-delimited protein table.
#' @param ctrl_cols   Character vector of control NSAF column names (n=3).
#' @param treat_cols  Character vector of treatment NSAF column names (n=3).
#' @param go_list_dirs Character vector of metaGOmics2 list_XXX directories.
#' @param ctrl_label  Short label for control condition.
#' @param treat_label Short label for treatment condition.
#' @param title       Plot title.
#' @param out_dir     Directory to save PNG + PDF outputs (created if needed).
#' @param out_prefix  File prefix for saved figures.
#' @return Invisible list: $data (annotated df), $plot (ggplot).
run_volcano <- function(nsaf_path,
                        ctrl_cols,
                        treat_cols,
                        go_list_dirs,
                        ctrl_label  = "Control",
                        treat_label = "Treatment",
                        title       = "Volcano",
                        out_dir     = "figures/volcano_plots",
                        out_prefix  = "volcano") {

  # ── 1. Load & filter ────────────────────────────────────────────────────────
  df <- read_tsv(nsaf_path, show_col_types = FALSE, na = c("", "NA")) %>%
    filter(!grepl("CONTAMINANT", `Protein Description(s)`,
                  ignore.case = TRUE, perl = TRUE)) %>%
    mutate(across(all_of(c(ctrl_cols, treat_cols)),
                  ~ as.numeric(.)))

  # Keep proteins detected (NSAF > 0) in ALL replicates of BOTH conditions
  df <- df %>%
    filter(
      rowSums(select(., all_of(ctrl_cols))  > 0) == length(ctrl_cols),
      rowSums(select(., all_of(treat_cols)) > 0) == length(treat_cols)
    )
  message(title, ": n=", nrow(df), " proteins (all replicates detected)")

  # ── 2. GO annotations ───────────────────────────────────────────────────────
  go_map <- load_go_map(go_list_dirs)

  cats <- lapply(df$`Protein(s)`, assign_go_category, prot_go_map = go_map)
  df$category <- vapply(cats, `[[`, character(1), "name")
  df$color    <- vapply(cats, `[[`, character(1), "color")

  # ── 3. Statistics ────────────────────────────────────────────────────────────
  # Welch's t-test on log10 NSAF; log2FC from raw NSAF means
  log_mat <- log10(as.matrix(df[, c(ctrl_cols, treat_cols)]))

  n_ctrl  <- length(ctrl_cols)
  n_treat <- length(treat_cols)

  stats <- lapply(seq_len(nrow(df)), function(i) {
    ctrl_vals  <- log_mat[i, seq_len(n_ctrl)]
    treat_vals <- log_mat[i, n_ctrl + seq_len(n_treat)]
    tt <- tryCatch(
      t.test(treat_vals, ctrl_vals, var.equal = FALSE),
      error = function(e) list(p.value = 1)
    )
    list(
      p_val   = if (is.nan(tt$p.value)) 1 else tt$p.value,
      log2FC  = log2(mean(as.numeric(df[i, treat_cols])) /
                     mean(as.numeric(df[i, ctrl_cols])))
    )
  })

  df$p_val  <- vapply(stats, `[[`, numeric(1), "p_val")
  df$log2FC <- vapply(stats, `[[`, numeric(1), "log2FC")
  df$FDR    <- p.adjust(df$p_val, method = "BH")

  # Y-axis: -log10(p) capped at 6
  y_cap         <- 6.0
  p_line        <- -log10(0.05)   # ≈ 1.301
  df$neg_log_p  <- pmin(-log10(pmax(df$p_val, 1e-10)), y_cap)

  df$sig     <- df$FDR < 0.05 & abs(df$log2FC) > 1
  df$up      <- df$sig & df$log2FC > 0
  df$down    <- df$sig & df$log2FC < 0

  message("  Sig (FDR<0.05 & |FC|>1): ", sum(df$sig),
          "  up=", sum(df$up), "  down=", sum(df$down))

  # ── 4. Build ordered category factor for legend ──────────────────────────────
  all_cat_names   <- c(vapply(GO_CATS, `[[`, character(1), "name"),
                       CAT_OTHER$name, CAT_HYPOTHETICAL$name)
  all_cat_colors  <- c(vapply(GO_CATS, `[[`, character(1), "color"),
                       CAT_OTHER$color, CAT_HYPOTHETICAL$color)
  names(all_cat_colors) <- all_cat_names

  # Only include categories present among significant points
  sig_cats <- intersect(all_cat_names, unique(df$category[df$sig]))
  legend_colors <- c("Non-significant" = "lightgray",
                     all_cat_colors[sig_cats])

  df$plot_cat <- ifelse(df$sig,
                        df$category,
                        "Non-significant")
  df$plot_cat <- factor(df$plot_cat,
                        levels = c("Non-significant", sig_cats))

  # ── 5. Hyperbolic boundary curves ────────────────────────────────────────────
  k     <- 1.0 * p_line
  xs_r  <- seq(1.001, 2.5, length.out = 400)
  xs_l  <- seq(-2.5, -1.001, length.out = 400)
  ys_r  <- k / (xs_r - 1)
  ys_l  <- k / (-xs_l - 1)

  curve_r <- data.frame(x = xs_r, y = ys_r) %>%
    filter(y > p_line, y <= y_cap)
  curve_l <- data.frame(x = xs_l, y = ys_l) %>%
    filter(y > p_line, y <= y_cap)

  # ── 6. Top 15 labels ─────────────────────────────────────────────────────────
  top15 <- df %>%
    filter(sig) %>%
    arrange(FDR) %>%
    head(15) %>%
    mutate(label = vapply(`Protein Description(s)`, clean_label, character(1)))

  # ── 7. Build plot ─────────────────────────────────────────────────────────────
  # Non-significant layer plotted first (gray, smaller)
  p <- ggplot() +
    geom_point(
      data = df %>% filter(!sig),
      aes(x = log2FC, y = neg_log_p),
      color = "lightgray", size = 1.0, alpha = 0.45, stroke = 0
    ) +
    # Significant points colored by GO category
    geom_point(
      data = df %>% filter(sig),
      aes(x = log2FC, y = neg_log_p, color = plot_cat),
      size = 2.2, alpha = 0.9, stroke = 0
    ) +
    # Threshold lines
    geom_hline(yintercept = p_line, linetype = "dashed",
               color = "black", linewidth = 0.5, alpha = 0.6) +
    geom_vline(xintercept =  1, linetype = "dashed",
               color = "black", linewidth = 0.5, alpha = 0.6) +
    geom_vline(xintercept = -1, linetype = "dashed",
               color = "black", linewidth = 0.5, alpha = 0.6) +
    # Hyperbolic boundary curves
    geom_line(data = curve_r, aes(x = x, y = y),
              color = "black", linewidth = 0.7) +
    geom_line(data = curve_l, aes(x = x, y = y),
              color = "black", linewidth = 0.7) +
    # Protein labels (top 15 by FDR)
    ggrepel::geom_text_repel(
      data         = top15,
      aes(x = log2FC, y = neg_log_p, label = label),
      size         = 2.2,
      color        = "black",
      segment.color = "gray60",
      segment.size  = 0.3,
      max.overlaps  = 20,
      box.padding   = 0.2,
      point.padding = 0.1
    ) +
    # Color scale for significant points
    scale_color_manual(
      values = legend_colors,
      name   = "GO biological process",
      guide  = guide_legend(
        override.aes = list(size = 3, alpha = 1),
        title.theme  = element_text(size = 8, face = "bold")
      )
    ) +
    # Axis limits
    coord_cartesian(xlim = c(-2.5, 2.5), ylim = c(0, y_cap)) +
    labs(
      title    = title,
      subtitle = sprintf(
        "n = %d proteins  ·  ↑ %d  ↓ %d significant (FDR < 0.05, |log₂FC| > 1)",
        nrow(df), sum(df$up), sum(df$down)
      ),
      x = bquote(log[2] ~ "NSAF  (" ~ .(treat_label) ~ "−" ~ .(ctrl_label) ~ ")"),
      y = expression(textstyle(-log[10] ~ p))
    ) +
    theme_bw(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold", size = 12),
      plot.subtitle    = element_text(color = "gray30", size = 9),
      legend.position  = "right",
      legend.text      = element_text(size = 7.5),
      legend.key.size  = unit(0.45, "cm"),
      axis.title       = element_text(size = 11)
    )

  # ── 8. Save ──────────────────────────────────────────────────────────────────
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  ggsave(file.path(out_dir, paste0(out_prefix, ".png")),
         p, width = 8, height = 7, dpi = 600, bg = "white")
  ggsave(file.path(out_dir, paste0(out_prefix, ".pdf")),
         p, width = 8, height = 7)
  message("  Saved to: ", out_dir)

  invisible(list(data = df, plot = p))
}
```

---

# 1 · CP34H — B-CTL vs B-1.5

*Colwellia psychrerythraea* 34H: control (HR_BC1–3) vs DNA Shield 1:1.5 (HR-E1–E3).
metaGOmics2 lists used: `list_000` and `list_001`.

```{r bact-volcano, fig.cap="CP34H: B-CTL vs B-1.5 GO-annotated volcano plot"}
bact_nsaf_path <- "data/Bacteria/Bactera_Comet_all_samples_proteins.txt"

bact_ctrl_cols  <- c("NSAF (HR_BC1)", "NSAF (HR_BC2)", "NSAF (HR_BC3)")
bact_treat_cols <- c("NSAF (HR-E1)",  "NSAF (HR_E2)",  "NSAF (HR_E3)")

bact_go_dirs <- c("data/metagomics2_results/list_000",
                  "data/metagomics2_results/list_001")

bact_result <- run_volcano(
  nsaf_path    = bact_nsaf_path,
  ctrl_cols    = bact_ctrl_cols,
  treat_cols   = bact_treat_cols,
  go_list_dirs = bact_go_dirs,
  ctrl_label   = "B-CTL",
  treat_label  = "B-1.5",
  title        = "B-CTL vs B-1.5 (Colwellia psychrerythraea 34H)",
  out_dir      = "figures/volcano_plots",
  out_prefix   = "bact_volcano_GO"
)

bact_result$plot
```

## CP34H significant proteins

```{r bact-sig-table}
bact_result$data %>%
  filter(sig) %>%
  arrange(FDR) %>%
  select(`Protein(s)`, `Protein Description(s)`, category,
         log2FC, p_val, FDR) %>%
  mutate(
    log2FC = round(log2FC, 3),
    p_val  = signif(p_val, 3),
    FDR    = signif(FDR,   3)
  ) %>%
  knitr::kable(
    caption = "Significant proteins: B-CTL vs B-1.5 (FDR < 0.05, |log2FC| > 1)"
  )
```

---

# 2 · T. pseudonana — P-CTL vs P-2.0

*Thalassiosira pseudonana*: control (PEC1–3) vs DNA Shield 1:2 (PED4–6 1:2).
metaGOmics2 lists used: `list_004` and `list_006`.

```{r phyto-volcano, fig.cap="T. pseudonana: P-CTL vs P-2.0 GO-annotated volcano plot"}
phyto_nsaf_path <- "data/Phytoplankton/Phytoplankton_Comet_all_samples_proteins.txt"

phyto_ctrl_cols  <- c("NSAF (PEC1)",     "NSAF (PEC2)",     "NSAF (PEC3)")
phyto_treat_cols <- c("NSAF (PED4 1:2)", "NSAF (PED5 1:2)", "NSAF (PED6 1:2)")

phyto_go_dirs <- c("data/metagomics2_results/list_004",
                   "data/metagomics2_results/list_006")

phyto_result <- run_volcano(
  nsaf_path    = phyto_nsaf_path,
  ctrl_cols    = phyto_ctrl_cols,
  treat_cols   = phyto_treat_cols,
  go_list_dirs = phyto_go_dirs,
  ctrl_label   = "P-CTL",
  treat_label  = "P-2.0",
  title        = "P-CTL vs P-2.0 (Thalassiosira pseudonana)",
  out_dir      = "figures/volcano_plots",
  out_prefix   = "phyto_volcano_GO"
)

phyto_result$plot
```

## T. pseudonana significant proteins

```{r phyto-sig-table}
phyto_result$data %>%
  filter(sig) %>%
  arrange(FDR) %>%
  select(`Protein(s)`, `Protein Description(s)`, category,
         log2FC, p_val, FDR) %>%
  mutate(
    log2FC = round(log2FC, 3),
    p_val  = signif(p_val, 3),
    FDR    = signif(FDR,   3)
  ) %>%
  knitr::kable(
    caption = "Significant proteins: P-CTL vs P-2.0 (FDR < 0.05, |log2FC| > 1)"
  )
```

---

# 3 · GO category breakdown

Summary of how many significant proteins fall into each GO category, across
both comparisons.

```{r go-summary}
# Combine both result sets
combined_sig <- bind_rows(
  bact_result$data  %>% filter(sig) %>% mutate(comparison = "B-CTL vs B-1.5"),
  phyto_result$data %>% filter(sig) %>% mutate(comparison = "P-CTL vs P-2.0")
)

combined_sig %>%
  count(comparison, category) %>%
  arrange(comparison, desc(n)) %>%
  knitr::kable(caption = "Significant proteins by GO category and comparison")
```

---

# Session info

```{r session}
sessionInfo()
```
